# Algebra of Matrices {#sec-matrices}

## Basic definitions

### Matrix notation

Let $\mathbb{K}$ be a field and $m, n \in \mathbb{N}$. An $m \times n$ matrix $A$ over the field $\mathbb{K}$ is a rectangular array of elements $a_{ij} \in \mathbb{K}$ arranged in $m$ rows and $n$ columns, usually presented in the following form:

$$A = [a_{ij}] = \begin{bmatrix}
a_{11} & a_{12} & \cdots & a_{1n} \\
a_{21} & a_{22} & \cdots & a_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
a_{m1} & a_{m2} & \cdots & a_{mn}
\end{bmatrix}$$

where $a_{ij}$ denotes the $(ij)$-th element (or entry) of the matrix, corresponding to row $i$ and column $j$.

**Matrix dimensions (or size)** refer to the number of rows and columns of a matrix, denoted as $m \times n$. The matrix $A$ is often written as $A_{m \times n}$ to make its dimensions explicit.


::: content-box-white
**NOTE:** While $\mathbb{K}$ usually represents the real numbers ($\mathbb{R}$) or complex numbers ($\mathbb{C}$), it can also represent other fields.
:::


The $m$ rows of matrix $A$ are $1 \times n$ matrices, also called **row vectors**:

$r_1 = \begin{bmatrix} a_{11} & a_{12} & ... & a_{1n} \end{bmatrix}$

$r_2 = \begin{bmatrix} a_{21} & a_{22} & ... & a_{2n} \end{bmatrix}$

$\vdots$

$r_m = \begin{bmatrix} a_{m1} & a_{m2} & ... & a_{mn} \end{bmatrix}$

 

The $n$ columns of $A$ are $m \times 1$ matrices, are called **column vectors**:

$c_1 = \begin{bmatrix} a_{11} \\ a_{21} \\ \vdots \\ a_{m1} \end{bmatrix}$, $c_2 = \begin{bmatrix} a_{12} \\ a_{22} \\ \vdots \\ a_{m2} \end{bmatrix}$, ..., $c_n = \begin{bmatrix} a_{1n} \\ a_{2n} \\ \vdots \\ a_{mn} \end{bmatrix}$

 

#### Example 1

The matrix 

$$A= \begin{bmatrix}
    4 & 0 & 2 & 1\\
    3 & 1 & 4 & 2\\
    2 & 0 & 1 & 3
  \end{bmatrix}$$

is a $3 \times 4$ matrix over the field of real numbers ($\mathbb{R}$). Its first row vector and first column vector are, respectively:

$$
r_1 = \begin{bmatrix} 4 & 0 & 2 & 1 \end{bmatrix}, \qquad c_1 = \begin{bmatrix} 4 \\ 3 \\ 2 \end{bmatrix}
$$


##### *Implementation in R*

We construct $A$ in R using the `matrix()` function. The entries of the matrix are first stored in a numeric vector named `entries` and then reshaped into a $3 \times 4$ matrix. Since the values in the vector are listed in row-wise order, we set the parameter `byrow = TRUE` which fills the matrix one row at a time before moving to the next. This ensures that the computational representation of A in R matches its mathematical form exactly.


```{r}
entries <- c(4, 0, 2, 1,
             3, 1, 4, 2,
             2, 0, 1, 3)

A <- matrix(entries, nrow = 3, ncol = 4, byrow = TRUE)
A
```


To verify the structure of the resulting object, we inspect its class and dimensions:

```{r}
class(A)
dim(A)
```

The function `class(A)` confirms that A is a matrix object, while `dim(A)` returns its dimensions, $3 \times 4$, consistent with the mathematical definition.

Matrix indexing in R follows the standard form `[row, column]`, allowing individual rows and columns to be extracted directly. By default, R drops the two-dimensional structure of the result. To prevent this and preserve the matrix form, the argument `drop = FALSE` should be used.

To extract the first row vector $r_1$, we use:

```{r}
r1 <- A[1, , drop = FALSE]
r1
class(r1)
dim(r1)
```



Similarly, the first column vector $c_1$ is obtained by:

```{r}
c1 <- A[, 1, drop = FALSE]
c1
class(c1)
dim(c1)
```




