# The Basics of Matrices {#sec-matrices}

## Matrix Notation

Let $\mathbb{K}$ be a field and $m, n \in \mathbb{N}$. An **$\mathbf{m} \times \mathbf{n}$ matrix** $A$ over the field $\mathbb{K}$ is a rectangular array of elements $a_{ij} \in \mathbb{K}$ arranged in $m$ rows and $n$ columns, usually presented in the following form:

$$A = [a_{ij}] = \begin{bmatrix}
a_{11} & a_{12} & \cdots & a_{1n} \\
a_{21} & a_{22} & \cdots & a_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
a_{m1} & a_{m2} & \cdots & a_{mn}
\end{bmatrix}$$

where $a_{ij}$ denotes the $(ij)$-th entry (or element) of the matrix, corresponding to row $i$ and column $j$.

**Matrix dimensions (or size)** refer to the number of rows and columns of a matrix, denoted as $m \times n$. The matrix $A$ is often written as $A_{m \times n}$ to make its dimensions explicit.


::: content-box-white
**NOTE:** While $\mathbb{K}$ usually represents the real numbers ($\mathbb{R}$) or complex numbers ($\mathbb{C}$), it can also represent other fields.
:::


The $m$ rows of matrix $A$ are $1 \times n$ matrices, also called **row vectors**:

$r_1 = \begin{bmatrix} a_{11} & a_{12} & ... & a_{1n} \end{bmatrix}$

$r_2 = \begin{bmatrix} a_{21} & a_{22} & ... & a_{2n} \end{bmatrix}$

$\vdots$

$r_m = \begin{bmatrix} a_{m1} & a_{m2} & ... & a_{mn} \end{bmatrix}$

 

The $n$ columns of $A$ are $m \times 1$ matrices, are called **column vectors**:

$c_1 = \begin{bmatrix} a_{11} \\ a_{21} \\ \vdots \\ a_{m1} \end{bmatrix}$, $c_2 = \begin{bmatrix} a_{12} \\ a_{22} \\ \vdots \\ a_{m2} \end{bmatrix}$, ..., $c_n = \begin{bmatrix} a_{1n} \\ a_{2n} \\ \vdots \\ a_{mn} \end{bmatrix}$

 

#### Example: A $3 \times 4$ Matrix

The matrix 

$$A_{3 \times 4}= \begin{bmatrix}
    4 & 0 & 2 & 1\\
    3 & 1 & 4 & 2\\
    2 & 0 & 1 & 3
  \end{bmatrix}$$

is a $3 \times 4$ matrix over the field of real numbers ($\mathbb{R}$). Its first row vector and first column vector are, respectively:

$$
r_1 = \begin{bmatrix} 4 & 0 & 2 & 1 \end{bmatrix}, \qquad c_1 = \begin{bmatrix} 4 \\ 3 \\ 2 \end{bmatrix}
$$


##### *Implementation in R*

We construct $A$ in R using the `matrix()` function. The entries of the matrix are first stored in a numeric vector named `entries` and then reshaped into a $3 \times 4$ matrix. Since the values in the vector are listed in row-wise order, we set the parameter `byrow = TRUE` which fills the matrix one row at a time before moving to the next. This ensures that the computational representation of A in R matches its mathematical form exactly.


```{r}
# Define entries of sigma_1 in row-wise order
entries <- c(4, 0, 2, 1,
             3, 1, 4, 2,
             2, 0, 1, 3)

# Construct 3x4 matrix using row-wise filling
A <- matrix(entries, nrow = 3, ncol = 4, byrow = TRUE)
A
```


To verify the structure of the resulting object, we inspect its class and dimensions:

```{r}
class(A)
dim(A)
```

The function `class(A)` confirms that A is a matrix object, while `dim(A)` returns its dimensions, $3 \times 4$, consistent with the mathematical definition.

Matrix indexing in R follows the standard form `[row, column]`, allowing individual rows and columns to be extracted directly. By default, R drops the two-dimensional structure of the result. To prevent this and preserve the matrix form, the argument `drop = FALSE` should be used.

To extract the first row vector $r_1$, we use:

```{r}
r1 <- A[1, , drop = FALSE]
r1
class(r1)
dim(r1)
```



Similarly, the first column vector $c_1$ is obtained by:

```{r}
c1 <- A[, 1, drop = FALSE]
c1
class(c1)
dim(c1)
```

 

#### Example: The Pauli Matrices

The Pauli matrices are three $2 \times 2$ complex matrices with fundamental applications in physics, particularly in quantum mechanics.

Let $\mathbb{K} = \mathbb{C}$ (the field of complex numbers). The Pauli matrices are defined as:

$$\sigma_1 = \begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix}, \quad 
\sigma_2 = \begin{bmatrix} 0 & -i \\ i & 0 \end{bmatrix}, \quad 
\sigma_3 = \begin{bmatrix} 1 & 0 \\ 0 & -1 \end{bmatrix}$$

 

##### *Implementation in R*

To represent these in R, we must account for the imaginary unit $i$, which is denoted as $i$ in R (e.g., $0+1i$). Because these matrices are $2 \times 2$, we set the dimensions accordingly.

- Matrix $\sigma_1$

```{r}
# Define entries of sigma_1 in row-wise order
entries <- c(0, 1,
             1, 0)

# Construct 2x2 matrix using row-wise filling
sigma_1 <- matrix(entries, nrow = 2, byrow = TRUE)
sigma_1
```

- Matrix $\sigma_2$

```{r}
# Use R's complex notation '1i' for the imaginary unit
entries <- c(0, 0-1i,
             0+1i, 0)

# Construct 2x2 matrix using row-wise filling
sigma_2 <- matrix(entries, nrow = 2, byrow = TRUE)
sigma_2
```

- Matrix $\sigma_3$

```{r}
# Define entries of sigma_2 in row-wise order
entries <- c(1, 0, 
             0, -1)

sigma_3 <- matrix(entries, nrow = 2, byrow = TRUE)
sigma_3
```

 

#### Example: The Planar Rotation Matrix

When matrix entries are functions, the matrix is called a matrix-valued function. A fundamental example in linear algebra and geometry is the planar rotation matrix $R(\theta)$, which represents a counter-clockwise rotation by an angle $\theta$ in $\mathbb{R}^2$. Mathematically, it is defined as:


$$R(\theta) = \begin{bmatrix} \cos(\theta) & -\sin(\theta) \\ \sin(\theta) & \cos(\theta) \end{bmatrix}$$
For $\theta = \frac{\pi}{4}$ (a rotation of $45^\circ$):

$$R\left(\frac{\pi}{4}\right) = 
\begin{bmatrix} 
\cos\left(\frac{\pi}{4}\right) & -\sin\left(\frac{\pi}{4}\right) \\ 
\sin\left(\frac{\pi}{4}\right) & \cos\left(\frac{\pi}{4}\right) 
\end{bmatrix} = 
\begin{bmatrix} 
\frac{\sqrt{2}}{2} & -\frac{\sqrt{2}}{2} \\ 
\frac{\sqrt{2}}{2} & \frac{\sqrt{2}}{2} 
\end{bmatrix}$$

 

##### *Implementation in R*

```{r}
# Define the matrix-valued function
R <- function(theta) {
  entries <- c(cos(theta), -sin(theta),
               sin(theta),  cos(theta))
  
  matrix(entries, nrow = 2, byrow = TRUE)
}

```


```{r}
# Define the angle of 45 degrees (pi/4 in radians)
theta <- pi / 4   

# Use the function to construct the matrix
R(theta)
```

 

The geometric effect of the planar rotation matrix $R(\theta)$ on a unit square for $\theta = \pi/4$ is shown in @fig-rotation.

```{r}
#| echo: false
#| warning: false
#| fig-align: center
#| label: fig-rotation
#| fig-cap: 'Planar rotation of the unit square by $R(\theta)$ around the origin $(0,0)$. At $\theta = \pi/4$, the square transforms into a diamond configuration: the pivot $\Sigma_1(0,0)$ remains invariant ($\Sigma_1 = \Sigma_1^\prime$), while the remaining vertices map to $\Sigma_2^\prime(\sqrt{2}/2, \sqrt{2}/2)$, the apex $\Sigma_3^\prime(0, \sqrt{2})$, and $\Sigma_4^\prime(-\sqrt{2}/2, \sqrt{2}/2)$.'

library(ggplot2)
library(dplyr)


rot_matrix <- R(theta)

# Define the unit square vertices
square_coords <- matrix(c(0,0, 1,0, 1,1, 0,1, 0,0), ncol = 2, byrow = TRUE)

# Apply rotation 
rotated_coords <- square_coords %*% t(rot_matrix)

# Data for polygons
plot_data <- data.frame(
  x = c(square_coords[,1], rotated_coords[,1]),
  y = c(square_coords[,2], rotated_coords[,2]),
  Shape = rep(c("Original", "Rotated"), each = 5)
)

# Data for the rotation arc
arc_data <- data.frame(
  x = 0.4 * cos(seq(0, theta, length.out = 50)),
  y = 0.4 * sin(seq(0, theta, length.out = 50))
)

# 7. ggplot Implementation
ggplot() +
  # --- AXIS ARROWS ---
  # X-axis arrow
  geom_segment(aes(x = -1.2, y = 0, xend = 1.6, yend = 0), 
               arrow = arrow(length = unit(0.2, "cm")), color = "black", size = 0.5) +
  # Y-axis arrow
  geom_segment(aes(x = 0, y = -0.2, xend = 0, yend = 1.7), 
               arrow = arrow(length = unit(0.2, "cm")), color = "black", size = 0.5) +
  # Axis labels
  annotate("text", x = 1.65, y = 0, label = "x", size = 5, fontface = "italic", hjust = 0) +
  annotate("text", x = 0, y = 1.75, label = "y", size = 5, fontface = "italic", vjust = 0) +

  # Draw the polygons
  geom_polygon(data = plot_data, aes(x = x, y = y, fill = Shape, color = Shape), alpha = 0.3, size = 1) +
  
  # --- THE STATIONARY ORIGIN ---
  # Sigma 1 = Sigma' 1: (0, 0)
  annotate("point", x = 0, y = 0, color = "gray40", size = 3) +
  annotate("text", x = -0.1, y = -0.15, 
           label = "Sigma[1] == Sigma[1]*minute ~ group('(', list(0, 0), ')')", 
           parse = TRUE, size = 4, color = "gray40", fontface = "bold", hjust = 1) +
  
  # --- ORIGINAL SQUARE CORNERS (Gray) ---
  # Sigma 2: (1, 0)
  annotate("point", x = 1, y = 0, color = "gray40", size = 3) +
  annotate("text", x = 1.05, y = -0.1, label = "Sigma[2] ~ group('(', list(1, 0), ')')", parse = TRUE, size = 4, color = "gray40", hjust = 0) +
  
  # Sigma 3: (1, 1)
  annotate("point", x = 1, y = 1, color = "gray40", size = 3) +
  annotate("text", x = 1.05, y = 1.05, label = "Sigma[3] ~ group('(', list(1, 1), ')')", parse = TRUE, size = 4, color = "gray40", hjust = 0) +
  
  # Sigma 4: (0, 1)
  annotate("point", x = 0, y = 1, color = "gray40", size = 3) +
  annotate("text", x = 0.02, y = 1.07, label = "Sigma[4] ~ group('(', list(0, 1), ')')", parse = TRUE, size = 4, color = "gray40", hjust = 1) +
  
  # --- ROTATED SQUARE CORNERS (Red) ---
  # Sigma' 2: (sqrt(2)/2, sqrt(2)/2)
  annotate("point", x = cos(theta), y = sin(theta), color = "darkred", size = 3) +
  annotate("text", x = cos(theta) + 0.05, y = sin(theta) - 0.05, 
           label = "Sigma[2]*minute ~ group('(', list(sqrt(2)/2, sqrt(2)/2), ')')", 
           parse = TRUE, size = 4, color = "darkred", fontface = "bold", hjust = 0) +
  
  # Sigma' 3: (0, sqrt(2))
  annotate("point", x = 0, y = sqrt(2), color = "darkred", size = 3) +
  annotate("text", x = 0, y = sqrt(2) + 0.1, 
           label = "Sigma[3]*minute ~ group('(', list(0, sqrt(2)), ')')", 
           parse = TRUE, size = 4, color = "darkred", fontface = "bold") +

  # Sigma' 4: (-sqrt(2)/2, sqrt(2)/2)
  annotate("point", x = -sin(theta), y = cos(theta), color = "darkred", size = 3) +
  annotate("text", x = -sin(theta) - 0.05, y = cos(theta) - 0.05, 
           label = "Sigma[4]*minute ~ group('(', list(-sqrt(2)/2, sqrt(2)/2), ')')", 
           parse = TRUE, size = 4, color = "darkred", fontface = "bold", hjust = 1) +
  
  # --- ANCILLARY ELEMENTS ---
  # CCW rotation arc and angle label
  geom_path(data = arc_data, aes(x = x, y = y), 
            arrow = arrow(length = unit(0.2, "cm"), type = "closed"), color = "black") +
  annotate("text", x = 0.55, y = 0.2, label = "theta == pi/4", parse = TRUE, size = 5) +
  
  # Plot setup
  coord_fixed(xlim = c(-1.4, 1.9), ylim = c(-0.4, 1.9)) +
  scale_fill_manual(values = c("gray30", "firebrick")) +
  scale_color_manual(values = c("gray40", "firebrick")) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text = element_blank(), # Optional: hides the grid numbers if you only want arrows
        panel.grid.minor = element_blank(),
        legend.position = "none")

```

 
## Special Types of Matrices

### The Null Matrix

An $m \times n$ matrix is called a **null matrix** (or zero matrix) if all its entries are zero, and is typically denoted by $O_{m \times n}$ or simply $O$:


$$O_{m \times n} = \begin{bmatrix}
0 & 0 & \cdots & 0 \\
0 & 0 & \cdots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \cdots & 0
\end{bmatrix}$$


#### Example: A $3 \times 4$ Null Matrix

$$O_{3 \times 4} = \begin{bmatrix}
0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0
\end{bmatrix}$$


##### *Implementation in R*

A null matrix is easily generated by specifying its dimensions and filling all entries with zero.

```{r}
O <- matrix(0, nrow = 3, ncol = 4)
O
```

 

### The Square Matrix

A matrix $A$ is called a **square matrix** if $m = n$, that is, if it has an equal number of rows and columns. In this case, $A$ is said to be of **order** $n$, denoted $A_{n \times n}$ or simply $A_n$. 


$$A_{n \times n} = \begin{bmatrix}
\boldsymbol{a_{11}} & a_{12}              & \cdots & a_{1n} \\
a_{21}              & \boldsymbol{a_{22}} & \cdots & a_{2n} \\
\vdots              & \vdots              & \ddots & \vdots \\
a_{n1}              & a_{n2}              & \cdots & \boldsymbol{a_{nn}}
\end{bmatrix}$$

 

The **main diagonal** (or principal diagonal) of this matrix consists of the entries $a_{ij}$ for which the row index equals the column index, that is, $i = j$ (shown in bold in the matrix), denoted:

$$\text{diag}(A) = \{a_{11}, a_{22}, \dots, a_{nn}\}$$

All remaining entries, where $i \neq j$, are referred to as off-diagonal elements.

::: content-box-white
**REMARK:** Although the main diagonal is most naturally associated with square matrices, it is defined for any $m \times n$ rectangular matrix, consisting of the entries $a_{ii}$ for $i = 1, 2, \ldots, \min(m, n)$.

- **Case 1: Horizontal Rectangular Matrix ($m < n$)**

For a $3 \times 4$ matrix, where $\min(3, 4) = 3$, the main diagonal contains exactly three entries. The diagonal terminates before reaching the final column.

$$A_{3 \times 4} = \begin{bmatrix} \mathbf{4} & 0 & 2 & 1 \\ 3 & \mathbf{1} & 4 & 2 \\ 2 & 0 & \mathbf{1} & 3 \end{bmatrix}$$

In this instance, $\text{diag}(A) = \{4, 1, 1\}$.


- **Case 2: Vertical Rectangular Matrix ($m > n$)**

For a $4 \times 3$ matrix, where $\min(4, 3) = 3$, the diagonal also contains three entries. Here, the diagonal is exhausted before reaching the final row.

$$B_{4 \times 3} = \begin{bmatrix} \mathbf{2} & 1 & 0 \\ 3 & \mathbf{-1} & 4 \\ 1 & 2 & \mathbf{5} \\ 0 & 3 & 1 \end{bmatrix}$$

Here, $\text{diag}(B) = \{2, -1, 5\}$. Note that the fourth row contains no diagonal entry because the index $i=4$ has no corresponding column index $j=4$.
:::

 

The **trace** of a square matrix $A_{n \times n}$ is the sum of its main diagonal entries, denoted by $\text{tr}(A)$:

$$\text{tr}(A) = a_{11} + a_{22} + \cdots + a_{nn} = \sum_{i=1}^{n} a_{ii}$$

The trace is defined only for square matrices.

 

#### Example: A $3 \times 3$ Square Matrix

Consider the following square matrix of order 3:

$$A_{3 \times 3} = \begin{bmatrix} 
\mathbf{5} & 1 & 0 \\ 
3 & \mathbf{-1} & 2 \\ 
4 & 0 & \mathbf{-1}
\end{bmatrix}$$



The main diagonal entries (shown in bold) are $a_{11} = 5$, $a_{22} = -1$, and $a_{33} = -1$, giving:

$$\text{diag}(A) = \{5, -1, -1\}$$  

$$\text{tr}(A) = 5 + (-1) + (-1) = 3$$

 

##### *Implementation in R*

We construct $A$ in R and use the built-in `diag()` function to extract its main diagonal entries. 

```{r}
entries <- c(5, 1, 0,
             3, -1, 2,
             4, 0, -1)

A <- matrix(entries, nrow = 3, ncol = 3, byrow = TRUE)
A
```




```{r}
# Main diagonal
diag(A)
```



The trace is not available as a standalone function in base R, but is straightforwardly computed as a sum.

```{r}
# Trace
sum(diag(A))
```


 

### The Diagonal Matrix

A **diagonal matrix**, typically denoted as $D$, is a square matrix in which all off-diagonal entries are zero, that is, $a_{ij} = 0$ for all $i \neq j$:

$$D_{n \times n} = \begin{bmatrix} a_{11} & 0 & \dots & 0 \\ 
0 & a_{22} & \dots & 0 \\ 
\vdots & \vdots & \ddots & \vdots \\ 
0 & 0 & \dots & a_{nn} 
\end{bmatrix}$$

 

#### Example: A $3 \times 3$ Diagonal Matrix


$$D_{3\times 3} =
  \begin{bmatrix}
    4 & 0 & 0\\
    0 & -1 & 0\\
    0 & 0 & -5
  \end{bmatrix}$$


with $\text{diag}(D) = \{4, -1, -5\}$ and $\text{tr}(D) = 4 + (-1) + (-5) = -2$.

 

##### *Implementation in R*

In R, a diagonal matrix can be constructed using the `diag()` function. 


```{r}
entries <- c(4, -1, -5)

D <- diag(entries)
D
```



```{r}
# Main diagonal and trace
diag(D)      
sum(diag(D))
```


 

### The Identity Matrix

An **identity matrix**, typically denoted as $I$, is a square matrix whose entries satisfy the condition:

$$a_{ij} = \begin{cases} 1 & \text{if } i = j \\ 0 & \text{if } i \neq j \end{cases}$$ 

That is, it has ones on the main diagonal and zeros everywhere else. In explicit form, it is written as:

$$I_{n \times n} = \begin{bmatrix} 1 & 0 & \dots & 0 \\ 
0 & 1 & \dots & 0 \\ 
\vdots & \vdots & \ddots & \vdots \\ 
0 & 0 & \dots & 1 
\end{bmatrix}$$

Thus, $I_{n \times n}$ is a special case of a diagonal matrix where all diagonal entries equal to 1.

 

#### Example: A $3 \times 3$ Identity Matrix

The identity matrix of order 3 is written as:

$$I_{3 \times 3} =
  \begin{bmatrix}
    1 & 0 & 0\\
    0 & 1 & 0\\
    0 & 0 & 1
  \end{bmatrix}$$
  


 

##### *Implementation in R*

In R, the `diag()` function is used to generate identity matrices. When passed a single integer $n$, it returns an $n \times n$ identity matrix.

```{r}
I <- diag(3)
I
```


