# The Basics of Matrices {#sec-matrices}

## Matrix Notation

Let $\mathbb{K}$ be a field and $m, n \in \mathbb{N}$. An $m \times n$ matrix $A$ over the field $\mathbb{K}$ is a rectangular array of elements $a_{ij} \in \mathbb{K}$ arranged in $m$ rows and $n$ columns, usually presented in the following form:

$$A = [a_{ij}] = \begin{bmatrix}
a_{11} & a_{12} & \cdots & a_{1n} \\
a_{21} & a_{22} & \cdots & a_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
a_{m1} & a_{m2} & \cdots & a_{mn}
\end{bmatrix}$$

where $a_{ij}$ denotes the $(ij)$-th element (or entry) of the matrix, corresponding to row $i$ and column $j$.

**Matrix dimensions (or size)** refer to the number of rows and columns of a matrix, denoted as $m \times n$. The matrix $A$ is often written as $A_{m \times n}$ to make its dimensions explicit.


::: content-box-white
**NOTE:** While $\mathbb{K}$ usually represents the real numbers ($\mathbb{R}$) or complex numbers ($\mathbb{C}$), it can also represent other fields.
:::


The $m$ rows of matrix $A$ are $1 \times n$ matrices, also called **row vectors**:

$r_1 = \begin{bmatrix} a_{11} & a_{12} & ... & a_{1n} \end{bmatrix}$

$r_2 = \begin{bmatrix} a_{21} & a_{22} & ... & a_{2n} \end{bmatrix}$

$\vdots$

$r_m = \begin{bmatrix} a_{m1} & a_{m2} & ... & a_{mn} \end{bmatrix}$

 

The $n$ columns of $A$ are $m \times 1$ matrices, are called **column vectors**:

$c_1 = \begin{bmatrix} a_{11} \\ a_{21} \\ \vdots \\ a_{m1} \end{bmatrix}$, $c_2 = \begin{bmatrix} a_{12} \\ a_{22} \\ \vdots \\ a_{m2} \end{bmatrix}$, ..., $c_n = \begin{bmatrix} a_{1n} \\ a_{2n} \\ \vdots \\ a_{mn} \end{bmatrix}$

 

#### Example: A $3 \times 4$ Matrix

The matrix 

$$A= \begin{bmatrix}
    4 & 0 & 2 & 1\\
    3 & 1 & 4 & 2\\
    2 & 0 & 1 & 3
  \end{bmatrix}$$

is a $3 \times 4$ matrix over the field of real numbers ($\mathbb{R}$). Its first row vector and first column vector are, respectively:

$$
r_1 = \begin{bmatrix} 4 & 0 & 2 & 1 \end{bmatrix}, \qquad c_1 = \begin{bmatrix} 4 \\ 3 \\ 2 \end{bmatrix}
$$


##### *Implementation in R*

We construct $A$ in R using the `matrix()` function. The entries of the matrix are first stored in a numeric vector named `entries` and then reshaped into a $3 \times 4$ matrix. Since the values in the vector are listed in row-wise order, we set the parameter `byrow = TRUE` which fills the matrix one row at a time before moving to the next. This ensures that the computational representation of A in R matches its mathematical form exactly.


```{r}
# Define entries of sigma_1 in row-wise order
entries <- c(4, 0, 2, 1,
             3, 1, 4, 2,
             2, 0, 1, 3)

# Construct 3x4 matrix using row-wise filling
A <- matrix(entries, nrow = 3, ncol = 4, byrow = TRUE)
A
```


To verify the structure of the resulting object, we inspect its class and dimensions:

```{r}
class(A)
dim(A)
```

The function `class(A)` confirms that A is a matrix object, while `dim(A)` returns its dimensions, $3 \times 4$, consistent with the mathematical definition.

Matrix indexing in R follows the standard form `[row, column]`, allowing individual rows and columns to be extracted directly. By default, R drops the two-dimensional structure of the result. To prevent this and preserve the matrix form, the argument `drop = FALSE` should be used.

To extract the first row vector $r_1$, we use:

```{r}
r1 <- A[1, , drop = FALSE]
r1
class(r1)
dim(r1)
```



Similarly, the first column vector $c_1$ is obtained by:

```{r}
c1 <- A[, 1, drop = FALSE]
c1
class(c1)
dim(c1)
```

 

#### Example: The Pauli Matrices

The Pauli matrices are three $2 \times 2$ complex matrices with fundamental applications in physics, particularly in quantum mechanics.

Let $\mathbb{K} = \mathbb{C}$ (the field of complex numbers). The Pauli matrices are defined as:

$$\sigma_1 = \begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix}, \quad 
\sigma_2 = \begin{bmatrix} 0 & -i \\ i & 0 \end{bmatrix}, \quad 
\sigma_3 = \begin{bmatrix} 1 & 0 \\ 0 & -1 \end{bmatrix}$$

 

##### *Implementation in R*

To represent these in R, we must account for the imaginary unit $i$, which is denoted as $i$ in R (e.g., $0+1i$). Because these matrices are $2 \times 2$, we set the dimensions accordingly.

- Matrix $\sigma_1$

```{r}
# Define entries of sigma_1 in row-wise order
entries <- c(0, 1,
             1, 0)

# Construct 2x2 matrix using row-wise filling
sigma_1 <- matrix(entries, nrow = 2, byrow = TRUE)
sigma_1
```

- Matrix $\sigma_2$

```{r}
# Use R's complex notation '1i' for the imaginary unit
entries <- c(0, 0-1i,
             0+1i, 0)

# Construct 2x2 matrix using row-wise filling
sigma_2 <- matrix(entries, nrow = 2, byrow = TRUE)
sigma_2
```

- Matrix $\sigma_3$

```{r}
# Define entries of sigma_2 in row-wise order
entries <- c(1, 0, 
             0, -1)

sigma_3 <- matrix(entries, nrow = 2, byrow = TRUE)
sigma_3
```

 

#### Example: The Planar Rotation Matrix

When matrix entries are functions, the matrix is called a matrix-valued function. A fundamental example in linear algebra and geometry is the planar rotation matrix $M(\theta)$, which represents a counter-clockwise rotation by an angle $\theta$ in $\mathbb{R}^2$. Mathematically, it is defined as:


$$M(\theta) = \begin{bmatrix} \cos(\theta) & -\sin(\theta) \\ \sin(\theta) & \cos(\theta) \end{bmatrix}$$
For $\theta = \frac{\pi}{4}$ (a rotation of $45^\circ$):

$$M\left(\frac{\pi}{4}\right) = 
\begin{bmatrix} 
\cos\left(\frac{\pi}{4}\right) & -\sin\left(\frac{\pi}{4}\right) \\ 
\sin\left(\frac{\pi}{4}\right) & \cos\left(\frac{\pi}{4}\right) 
\end{bmatrix} = 
\begin{bmatrix} 
\frac{\sqrt{2}}{2} & -\frac{\sqrt{2}}{2} \\ 
\frac{\sqrt{2}}{2} & \frac{\sqrt{2}}{2} 
\end{bmatrix}$$

 

##### *Implementation in R*

```{r}
# Define the matrix-valued function
M <- function(theta) {
  entries <- c(cos(theta), -sin(theta),
               sin(theta),  cos(theta))
  
  matrix(entries, nrow = 2, byrow = TRUE)
}

```


```{r}
# Define the angle of 45 degrees (pi/4 in radians)
theta <- pi / 4   

# Use the function to construct the matrix
M(theta)
```

 

The geometric effect of the planar rotation matrix $M(\frac{\pi}{4})$ on a unit square is shown below.

```{r}
#| echo: false
#| warning: false
#| fig-align: center
#| label: fig-rotation
#| fig-cap: "Planar rotation of a unit square by $\\pi/4$."

library(ggplot2)
library(dplyr)


rot_matrix <- M(theta)

# Define the unit square vertices
square_coords <- matrix(c(0,0, 1,0, 1,1, 0,1, 0,0), ncol = 2, byrow = TRUE)

# Apply rotation 
rotated_coords <- square_coords %*% t(rot_matrix)

# Data for polygons
plot_data <- data.frame(
  x = c(square_coords[,1], rotated_coords[,1]),
  y = c(square_coords[,2], rotated_coords[,2]),
  Shape = rep(c("Original", "Rotated"), each = 5)
)

# Data for the rotation arc
arc_data <- data.frame(
  x = 0.4 * cos(seq(0, theta, length.out = 50)),
  y = 0.4 * sin(seq(0, theta, length.out = 50))
)

# 7. ggplot Implementation
ggplot() +
  # --- AXIS ARROWS ---
  # X-axis arrow
  geom_segment(aes(x = -1.2, y = 0, xend = 1.6, yend = 0), 
               arrow = arrow(length = unit(0.2, "cm")), color = "black", size = 0.5) +
  # Y-axis arrow
  geom_segment(aes(x = 0, y = -0.2, xend = 0, yend = 1.7), 
               arrow = arrow(length = unit(0.2, "cm")), color = "black", size = 0.5) +
  # Axis labels
  annotate("text", x = 1.65, y = 0, label = "x", size = 5, fontface = "italic", hjust = 0) +
  annotate("text", x = 0, y = 1.75, label = "y", size = 5, fontface = "italic", vjust = 0) +

  # Draw the polygons
  geom_polygon(data = plot_data, aes(x = x, y = y, fill = Shape, color = Shape), alpha = 0.3, size = 1) +
  
  # --- THE STATIONARY ORIGIN ---
  # Sigma 1 = Sigma' 1: (0, 0)
  annotate("point", x = 0, y = 0, color = "gray40", size = 3) +
  annotate("text", x = -0.1, y = -0.15, 
           label = "Sigma[1] == Sigma[1]*minute ~ group('(', list(0, 0), ')')", 
           parse = TRUE, size = 4, color = "gray40", fontface = "bold", hjust = 1) +
  
  # --- ORIGINAL SQUARE CORNERS (Gray) ---
  # Sigma 2: (1, 0)
  annotate("point", x = 1, y = 0, color = "gray40", size = 3) +
  annotate("text", x = 1.05, y = -0.1, label = "Sigma[2] ~ group('(', list(1, 0), ')')", parse = TRUE, size = 4, color = "gray40", hjust = 0) +
  
  # Sigma 3: (1, 1)
  annotate("point", x = 1, y = 1, color = "gray40", size = 3) +
  annotate("text", x = 1.05, y = 1.05, label = "Sigma[3] ~ group('(', list(1, 1), ')')", parse = TRUE, size = 4, color = "gray40", hjust = 0) +
  
  # Sigma 4: (0, 1)
  annotate("point", x = 0, y = 1, color = "gray40", size = 3) +
  annotate("text", x = 0.02, y = 1.07, label = "Sigma[4] ~ group('(', list(0, 1), ')')", parse = TRUE, size = 4, color = "gray40", hjust = 1) +
  
  # --- ROTATED SQUARE CORNERS (Red) ---
  # Sigma' 2: (sqrt(2)/2, sqrt(2)/2)
  annotate("point", x = cos(theta), y = sin(theta), color = "darkred", size = 3) +
  annotate("text", x = cos(theta) + 0.05, y = sin(theta) - 0.05, 
           label = "Sigma[2]*minute ~ group('(', list(sqrt(2)/2, sqrt(2)/2), ')')", 
           parse = TRUE, size = 4, color = "darkred", fontface = "bold", hjust = 0) +
  
  # Sigma' 3: (0, sqrt(2))
  annotate("point", x = 0, y = sqrt(2), color = "darkred", size = 3) +
  annotate("text", x = 0, y = sqrt(2) + 0.1, 
           label = "Sigma[3]*minute ~ group('(', list(0, sqrt(2)), ')')", 
           parse = TRUE, size = 4, color = "darkred", fontface = "bold") +

  # Sigma' 4: (-sqrt(2)/2, sqrt(2)/2)
  annotate("point", x = -sin(theta), y = cos(theta), color = "darkred", size = 3) +
  annotate("text", x = -sin(theta) - 0.05, y = cos(theta) - 0.05, 
           label = "Sigma[4]*minute ~ group('(', list(-sqrt(2)/2, sqrt(2)/2), ')')", 
           parse = TRUE, size = 4, color = "darkred", fontface = "bold", hjust = 1) +
  
  # --- ANCILLARY ELEMENTS ---
  # CCW rotation arc and angle label
  geom_path(data = arc_data, aes(x = x, y = y), 
            arrow = arrow(length = unit(0.2, "cm"), type = "closed"), color = "black") +
  annotate("text", x = 0.55, y = 0.2, label = "theta == pi/4", parse = TRUE, size = 5) +
  
  # Plot setup
  coord_fixed(xlim = c(-1.4, 1.9), ylim = c(-0.4, 1.9)) +
  scale_fill_manual(values = c("gray30", "firebrick")) +
  scale_color_manual(values = c("gray40", "firebrick")) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text = element_blank(), # Optional: hides the grid numbers if you only want arrows
        panel.grid.minor = element_blank(),
        legend.position = "none")

```


In @fig-rotation the planar rotation matrix $M(\theta)$ rotates the unit square counter-clockwise around the origin $(0,0)$. At $\theta = \pi/4$, the square maps to a diamond configuration, where the vertices are transformed as follows:
$\Sigma_1(0,0)$ remains fixed at $\Sigma'_1(0,0)$; $\Sigma_2(1,0)$ maps to $\Sigma'_2(\sqrt{2}/2, \sqrt{2}/2)$; $\Sigma_3(1,1)$ rotates to the apex $\Sigma'_3(0, \sqrt{2})$; and $\Sigma_4(0,1)$ maps to $\Sigma'_4(-\sqrt{2}/2, \sqrt{2}/2)$. 


## Special Types of Matrices

### The Null Matrix

An $m \times n$ matrix is called a null matrix (or zero matrix) if all its entries are zero, and is typically denoted by $O_{m \times n}$ or simply $O$:


$$A = \begin{bmatrix}
0 & 0 & \cdots & 0 \\
0 & 0 & \cdots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \cdots & 0
\end{bmatrix} = [a_{ij}] \equiv O$$


#### Example: A $3 \times 4$ Null Matrix

$$O_{3 \times 4} = \begin{bmatrix}
0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0
\end{bmatrix}$$


##### *Implementation in R*

A null matrix is easily generated by specifying its dimensions and filling all entries with zero.

```{r}
O <- matrix(0, nrow = 3, ncol = 4)
O
```

 

### The Square Matrix

If a matrix $A$ has $m$ rows and $n$ columns, it is called a square matrix if $m = n$, that is, if it has an equal number of rows and columns. In this case, $A$ is said to be of order $n$, denoted $A_{n \times n}$ or simply $A_n$.


$$A_{n \times n} = \begin{bmatrix}
a_{11} & a_{12} & \cdots & a_{1n} \\
a_{21} & a_{22} & \cdots & a_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
a_{n1} & a_{n2} & \cdots & a_{nn}
\end{bmatrix}$$


The most significant structural feature of a square matrix is the main diagonal (or principal diagonal). This diagonal consists of the set of entries $a_{ij}$ where the row index $i$ is equal to the column index $j$:$$\text{diag}(A) = \{a_{11}, a_{22}, \dots, a_{nn}\}$$Entries not located on this line are referred to as off-diagonal elements.



#### Example: A $3 \times 3$ Square Matrix


$$A_{3 \times 3} =
  \begin{bmatrix}
    5 & 1 & 0\\
    3 & -1 & 2\\
    4 & 0 & -1
    \end{bmatrix}$$




##### *Implementation in R*

```{r}
A <- matrix(c(5, 1, 0,
              3, -1, 2,
              4, 0, -1),
            nrow = 3, ncol = 3, byrow = TRUE)

A
```

