# The Basics of Matrices {#sec-matrices}

## Matrix Notation

Let $\mathbb{K}$ be a field and $m, n \in \mathbb{N}$. An $m \times n$ matrix $A$ over the field $\mathbb{K}$ is a rectangular array of elements $a_{ij} \in \mathbb{K}$ arranged in $m$ rows and $n$ columns, usually presented in the following form:

$$A = [a_{ij}] = \begin{bmatrix}
a_{11} & a_{12} & \cdots & a_{1n} \\
a_{21} & a_{22} & \cdots & a_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
a_{m1} & a_{m2} & \cdots & a_{mn}
\end{bmatrix}$$

where $a_{ij}$ denotes the $(ij)$-th element (or entry) of the matrix, corresponding to row $i$ and column $j$.

**Matrix dimensions (or size)** refer to the number of rows and columns of a matrix, denoted as $m \times n$. The matrix $A$ is often written as $A_{m \times n}$ to make its dimensions explicit.


::: content-box-white
**NOTE:** While $\mathbb{K}$ usually represents the real numbers ($\mathbb{R}$) or complex numbers ($\mathbb{C}$), it can also represent other fields.
:::


The $m$ rows of matrix $A$ are $1 \times n$ matrices, also called **row vectors**:

$r_1 = \begin{bmatrix} a_{11} & a_{12} & ... & a_{1n} \end{bmatrix}$

$r_2 = \begin{bmatrix} a_{21} & a_{22} & ... & a_{2n} \end{bmatrix}$

$\vdots$

$r_m = \begin{bmatrix} a_{m1} & a_{m2} & ... & a_{mn} \end{bmatrix}$

 

The $n$ columns of $A$ are $m \times 1$ matrices, are called **column vectors**:

$c_1 = \begin{bmatrix} a_{11} \\ a_{21} \\ \vdots \\ a_{m1} \end{bmatrix}$, $c_2 = \begin{bmatrix} a_{12} \\ a_{22} \\ \vdots \\ a_{m2} \end{bmatrix}$, ..., $c_n = \begin{bmatrix} a_{1n} \\ a_{2n} \\ \vdots \\ a_{mn} \end{bmatrix}$

 

#### Example 1

The matrix 

$$A= \begin{bmatrix}
    4 & 0 & 2 & 1\\
    3 & 1 & 4 & 2\\
    2 & 0 & 1 & 3
  \end{bmatrix}$$

is a $3 \times 4$ matrix over the field of real numbers ($\mathbb{R}$). Its first row vector and first column vector are, respectively:

$$
r_1 = \begin{bmatrix} 4 & 0 & 2 & 1 \end{bmatrix}, \qquad c_1 = \begin{bmatrix} 4 \\ 3 \\ 2 \end{bmatrix}
$$


##### *Implementation in R*

We construct $A$ in R using the `matrix()` function. The entries of the matrix are first stored in a numeric vector named `entries` and then reshaped into a $3 \times 4$ matrix. Since the values in the vector are listed in row-wise order, we set the parameter `byrow = TRUE` which fills the matrix one row at a time before moving to the next. This ensures that the computational representation of A in R matches its mathematical form exactly.


```{r}
# Define entries of sigma_1 in row-wise order
entries <- c(4, 0, 2, 1,
             3, 1, 4, 2,
             2, 0, 1, 3)

# Construct 3x4 matrix using row-wise filling
A <- matrix(entries, nrow = 3, ncol = 4, byrow = TRUE)
A
```


To verify the structure of the resulting object, we inspect its class and dimensions:

```{r}
class(A)
dim(A)
```

The function `class(A)` confirms that A is a matrix object, while `dim(A)` returns its dimensions, $3 \times 4$, consistent with the mathematical definition.

Matrix indexing in R follows the standard form `[row, column]`, allowing individual rows and columns to be extracted directly. By default, R drops the two-dimensional structure of the result. To prevent this and preserve the matrix form, the argument `drop = FALSE` should be used.

To extract the first row vector $r_1$, we use:

```{r}
r1 <- A[1, , drop = FALSE]
r1
class(r1)
dim(r1)
```



Similarly, the first column vector $c_1$ is obtained by:

```{r}
c1 <- A[, 1, drop = FALSE]
c1
class(c1)
dim(c1)
```

 

#### Example 2: The Pauli Matrices

The Pauli matrices are three $2 \times 2$ complex matrices with fundamental applications in physics, particularly in quantum mechanics.

Let $\mathbb{K} = \mathbb{C}$ (the field of complex numbers). The Pauli matrices are defined as:

$$\sigma_1 = \begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix}, \quad 
\sigma_2 = \begin{bmatrix} 0 & -i \\ i & 0 \end{bmatrix}, \quad 
\sigma_3 = \begin{bmatrix} 1 & 0 \\ 0 & -1 \end{bmatrix}$$

 

##### *Implementation in R*

To represent these in R, we must account for the imaginary unit $i$, which is denoted as $i$ in R (e.g., $0+1i$). Because these matrices are $2 \times 2$, we set the dimensions accordingly.

- Matrix $\sigma_1$

```{r}
# Define entries of sigma_1 in row-wise order
entries <- c(0, 1,
             1, 0)

# Construct 2x2 matrix using row-wise filling
sigma_1 <- matrix(entries, nrow = 2, byrow = TRUE)
sigma_1
```

- Matrix $\sigma_2$

```{r}
# Use R's complex notation '1i' for the imaginary unit
entries <- c(0, 0-1i,
             0+1i, 0)

# Construct 2x2 matrix using row-wise filling
sigma_2 <- matrix(entries, nrow = 2, byrow = TRUE)
sigma_2
```

- Matrix $\sigma_3$

```{r}
# Define entries of sigma_2 in row-wise order
entries <- c(1, 0, 
             0, -1)

sigma_3 <- matrix(entries, nrow = 2, byrow = TRUE)
sigma_3
```

 

#### Example 3: The Planar Rotation Matrix

When matrix entries are functions, the matrix is called a matrix-valued function. A fundamental example in linear algebra and geometry is the planar rotation matrix $M(\theta)$, which represents a counter-clockwise rotation by an angle $\theta$ in $\mathbb{R}^2$. Mathematically, it is defined as:


$$M(\theta) = \begin{bmatrix} \cos(\theta) & -\sin(\theta) \\ \sin(\theta) & \cos(\theta) \end{bmatrix}$$
For $\theta = \frac{\pi}{4}$ (a rotation of $45^\circ$):

$$M\left(\frac{\pi}{4}\right) = 
\begin{bmatrix} 
\cos\left(\frac{\pi}{4}\right) & -\sin\left(\frac{\pi}{4}\right) \\ 
\sin\left(\frac{\pi}{4}\right) & \cos\left(\frac{\pi}{4}\right) 
\end{bmatrix} = 
\begin{bmatrix} 
\frac{\sqrt{2}}{2} & -\frac{\sqrt{2}}{2} \\ 
\frac{\sqrt{2}}{2} & \frac{\sqrt{2}}{2} 
\end{bmatrix}$$

 

##### *Implementation in R*

```{r}
# Define the matrix-valued function
M <- function(theta) {
  entries <- c(cos(theta), -sin(theta),
               sin(theta),  cos(theta))
  
  matrix(entries, nrow = 2, byrow = TRUE)
}

```


```{r}
# Define the angle of 45 degrees (pi/4 in radians)
theta <- pi / 4   

# Use the function to construct the matrix
M(theta)
```



## Special Types of Matrices

### The Square Matrix

If a matrix $A$ has $m$ rows and $n$ columns, it is called a square matrix if $m = n$, that is, if it has an equal number of rows and columns. In this case, $A$ is said to be of order $n$, denoted $A_{n \times n}$ or simply $A_n$.



