# The Basics of Matrices {#sec-matrices}

## Matrix Notation

Let $\mathbb{K}$ be a field and $m, n \in \mathbb{N}$. An **$\mathbf{m} \times \mathbf{n}$ matrix** $A$ over the field $\mathbb{K}$ is a rectangular array of elements $a_{ij} \in \mathbb{K}$ arranged in $m$ rows and $n$ columns, usually presented in the following form:

$$A = [a_{ij}] = \begin{bmatrix}
a_{11} & a_{12} & \cdots & a_{1n} \\
a_{21} & a_{22} & \cdots & a_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
a_{m1} & a_{m2} & \cdots & a_{mn}
\end{bmatrix}$$

where $a_{ij}$ denotes the $(ij)$-th entry (or element) of the matrix, corresponding to row $i$ and column $j$.

**Matrix dimensions (or size)** refer to the number of rows and columns of a matrix, denoted as $m \times n$. The matrix $A$ is often written as $A_{m \times n}$ to make its dimensions explicit.


::: content-box-white
**NOTE:** While $\mathbb{K}$ usually represents the real numbers ($\mathbb{R}$) or complex numbers ($\mathbb{C}$), it can also represent other fields.
:::


The $m$ rows of matrix $A$ are $1 \times n$ matrices, also called **row vectors** (or row matrix):

$r_1 = \begin{bmatrix} a_{11} & a_{12} & ... & a_{1n} \end{bmatrix}$

$r_2 = \begin{bmatrix} a_{21} & a_{22} & ... & a_{2n} \end{bmatrix}$

$\vdots$

$r_m = \begin{bmatrix} a_{m1} & a_{m2} & ... & a_{mn} \end{bmatrix}$

 

The $n$ columns of $A$ are $m \times 1$ matrices, are called **column vectors** (or column matrix):

$c_1 = \begin{bmatrix} a_{11} \\ a_{21} \\ \vdots \\ a_{m1} \end{bmatrix}$, $c_2 = \begin{bmatrix} a_{12} \\ a_{22} \\ \vdots \\ a_{m2} \end{bmatrix}$, ..., $c_n = \begin{bmatrix} a_{1n} \\ a_{2n} \\ \vdots \\ a_{mn} \end{bmatrix}$

 

#### Example: A $3 \times 4$ Matrix

The matrix 

$$A_{3 \times 4}= \begin{bmatrix}
    4 & 0 & 2 & 1\\
    3 & 1 & 4 & 2\\
    2 & 0 & 1 & 3
  \end{bmatrix}$$

is a $3 \times 4$ rectangular matrix over the field of real numbers ($\mathbb{R}$). Its first row vector and first column vector are, respectively:

$$
r_1 = \begin{bmatrix} 4 & 0 & 2 & 1 \end{bmatrix}, \qquad c_1 = \begin{bmatrix} 4 \\ 3 \\ 2 \end{bmatrix}
$$


##### *Implementation in R*

We construct $A$ in R using the `matrix()` function. The entries of the matrix are first stored in a numeric vector named `entries` and then reshaped into a $3 \times 4$ matrix. Since the values in the vector are listed in row-wise order, we set the parameter `byrow = TRUE` which fills the matrix one row at a time before moving to the next. This ensures that the computational representation of A in R matches its mathematical form exactly.


```{r}
# Define entries of sigma_1 in row-wise order
entries <- c(4, 0, 2, 1,
             3, 1, 4, 2,
             2, 0, 1, 3)

# Construct 3x4 matrix using row-wise filling
A <- matrix(entries, nrow = 3, ncol = 4, byrow = TRUE)
A
```


To verify the structure of the resulting object, we inspect its class and dimensions:

```{r}
class(A)
dim(A)
```

The function `class(A)` confirms that A is a matrix object, while `dim(A)` returns its dimensions, $3 \times 4$, consistent with the mathematical definition.

Matrix indexing in R follows the standard form `[row, column]`, allowing individual rows and columns to be extracted directly. By default, R drops the two-dimensional structure of the result. To prevent this and preserve the matrix form, the argument `drop = FALSE` should be used.

To extract the first row vector $r_1$, we use:

```{r}
r1 <- A[1, , drop = FALSE]
r1
class(r1)
dim(r1)
```



Similarly, the first column vector $c_1$ is obtained by:

```{r}
c1 <- A[, 1, drop = FALSE]
c1
class(c1)
dim(c1)
```

 

#### Example: The Pauli Matrices

The Pauli matrices are three $2 \times 2$ complex matrices with fundamental applications in physics, particularly in quantum mechanics.

Let $\mathbb{K} = \mathbb{C}$ (the field of complex numbers). The Pauli matrices are defined as:

$$\sigma_1 = \begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix}, \quad 
\sigma_2 = \begin{bmatrix} 0 & -i \\ i & 0 \end{bmatrix}, \quad 
\sigma_3 = \begin{bmatrix} 1 & 0 \\ 0 & -1 \end{bmatrix}$$

 

##### *Implementation in R*

To represent these in R, we must account for the imaginary unit $i$, which is denoted as $i$ in R (e.g., $0+1i$). Because these matrices are $2 \times 2$, we set the dimensions accordingly.

- Matrix $\sigma_1$

```{r}
# Define entries of sigma_1 in row-wise order
entries <- c(0, 1,
             1, 0)

# Construct 2x2 matrix using row-wise filling
sigma_1 <- matrix(entries, nrow = 2, byrow = TRUE)
sigma_1
```

- Matrix $\sigma_2$

```{r}
# Use R's complex notation '1i' for the imaginary unit
entries <- c(0, 0-1i,
             0+1i, 0)

# Construct 2x2 matrix using row-wise filling
sigma_2 <- matrix(entries, nrow = 2, byrow = TRUE)
sigma_2
```

- Matrix $\sigma_3$

```{r}
# Define entries of sigma_2 in row-wise order
entries <- c(1, 0, 
             0, -1)

sigma_3 <- matrix(entries, nrow = 2, byrow = TRUE)
sigma_3
```

 

#### Example: The Planar Rotation Matrix

When matrix entries are functions, the matrix is called a matrix-valued function. A fundamental example in linear algebra and geometry is the planar rotation matrix $R(\theta)$, which represents a counter-clockwise rotation by an angle $\theta$ in $\mathbb{R}^2$. Mathematically, it is defined as:


$$R(\theta) = \begin{bmatrix} \cos(\theta) & -\sin(\theta) \\ \sin(\theta) & \cos(\theta) \end{bmatrix}$$
For $\theta = \frac{\pi}{4}$ (a rotation of $45^\circ$):

$$R\left(\frac{\pi}{4}\right) = 
\begin{bmatrix} 
\cos\left(\frac{\pi}{4}\right) & -\sin\left(\frac{\pi}{4}\right) \\ 
\sin\left(\frac{\pi}{4}\right) & \cos\left(\frac{\pi}{4}\right) 
\end{bmatrix} = 
\begin{bmatrix} 
\frac{\sqrt{2}}{2} & -\frac{\sqrt{2}}{2} \\ 
\frac{\sqrt{2}}{2} & \frac{\sqrt{2}}{2} 
\end{bmatrix}$$

 

##### *Implementation in R*

```{r}
# Define the matrix-valued function
R <- function(theta) {
  entries <- c(cos(theta), -sin(theta),
               sin(theta),  cos(theta))
  
  matrix(entries, nrow = 2, byrow = TRUE)
}

```


```{r}
# Define the angle of 45 degrees (pi/4 in radians)
theta <- pi / 4   

# Use the function to construct the matrix
R(theta)
```

 

The geometric effect of the planar rotation matrix $R(\theta)$ on a unit square for $\theta = \pi/4$ is shown in @fig-rotation.

```{r}
#| echo: false
#| warning: false
#| fig-align: center
#| label: fig-rotation
#| fig-cap: 'Planar rotation of the unit square by $R(\theta)$ around the origin $(0,0)$. At $\theta = \pi/4$, the square transforms into a diamond configuration: the pivot $\Sigma_1(0,0)$ remains invariant ($\Sigma_1 = \Sigma_1^\prime$), while the remaining vertices map to $\Sigma_2^\prime(\sqrt{2}/2, \sqrt{2}/2)$, the apex $\Sigma_3^\prime(0, \sqrt{2})$, and $\Sigma_4^\prime(-\sqrt{2}/2, \sqrt{2}/2)$.'

library(ggplot2)
library(dplyr)


rot_matrix <- R(theta)

# Define the unit square vertices
square_coords <- matrix(c(0,0, 1,0, 1,1, 0,1, 0,0), ncol = 2, byrow = TRUE)

# Apply rotation 
rotated_coords <- square_coords %*% t(rot_matrix)

# Data for polygons
plot_data <- data.frame(
  x = c(square_coords[,1], rotated_coords[,1]),
  y = c(square_coords[,2], rotated_coords[,2]),
  Shape = rep(c("Original", "Rotated"), each = 5)
)

# Data for the rotation arc
arc_data <- data.frame(
  x = 0.4 * cos(seq(0, theta, length.out = 50)),
  y = 0.4 * sin(seq(0, theta, length.out = 50))
)

# 7. ggplot Implementation
ggplot() +
  # --- AXIS ARROWS ---
  # X-axis arrow
  geom_segment(aes(x = -1.2, y = 0, xend = 1.6, yend = 0), 
               arrow = arrow(length = unit(0.2, "cm")), color = "black", size = 0.5) +
  # Y-axis arrow
  geom_segment(aes(x = 0, y = -0.2, xend = 0, yend = 1.7), 
               arrow = arrow(length = unit(0.2, "cm")), color = "black", size = 0.5) +
  # Axis labels
  annotate("text", x = 1.65, y = 0, label = "x", size = 5, fontface = "italic", hjust = 0) +
  annotate("text", x = 0, y = 1.75, label = "y", size = 5, fontface = "italic", vjust = 0) +

  # Draw the polygons
  geom_polygon(data = plot_data, aes(x = x, y = y, fill = Shape, color = Shape), alpha = 0.3, size = 1) +
  
  # --- THE STATIONARY ORIGIN ---
  # Sigma 1 = Sigma' 1: (0, 0)
  annotate("point", x = 0, y = 0, color = "gray40", size = 3) +
  annotate("text", x = -0.1, y = -0.15, 
           label = "Sigma[1] == Sigma[1]*minute ~ group('(', list(0, 0), ')')", 
           parse = TRUE, size = 4, color = "gray40", fontface = "bold", hjust = 1) +
  
  # --- ORIGINAL SQUARE CORNERS (Gray) ---
  # Sigma 2: (1, 0)
  annotate("point", x = 1, y = 0, color = "gray40", size = 3) +
  annotate("text", x = 1.05, y = -0.1, label = "Sigma[2] ~ group('(', list(1, 0), ')')", parse = TRUE, size = 4, color = "gray40", hjust = 0) +
  
  # Sigma 3: (1, 1)
  annotate("point", x = 1, y = 1, color = "gray40", size = 3) +
  annotate("text", x = 1.05, y = 1.05, label = "Sigma[3] ~ group('(', list(1, 1), ')')", parse = TRUE, size = 4, color = "gray40", hjust = 0) +
  
  # Sigma 4: (0, 1)
  annotate("point", x = 0, y = 1, color = "gray40", size = 3) +
  annotate("text", x = 0.02, y = 1.07, label = "Sigma[4] ~ group('(', list(0, 1), ')')", parse = TRUE, size = 4, color = "gray40", hjust = 1) +
  
  # --- ROTATED SQUARE CORNERS (Red) ---
  # Sigma' 2: (sqrt(2)/2, sqrt(2)/2)
  annotate("point", x = cos(theta), y = sin(theta), color = "darkred", size = 3) +
  annotate("text", x = cos(theta) + 0.05, y = sin(theta) - 0.05, 
           label = "Sigma[2]*minute ~ group('(', list(sqrt(2)/2, sqrt(2)/2), ')')", 
           parse = TRUE, size = 4, color = "darkred", fontface = "bold", hjust = 0) +
  
  # Sigma' 3: (0, sqrt(2))
  annotate("point", x = 0, y = sqrt(2), color = "darkred", size = 3) +
  annotate("text", x = 0, y = sqrt(2) + 0.1, 
           label = "Sigma[3]*minute ~ group('(', list(0, sqrt(2)), ')')", 
           parse = TRUE, size = 4, color = "darkred", fontface = "bold") +

  # Sigma' 4: (-sqrt(2)/2, sqrt(2)/2)
  annotate("point", x = -sin(theta), y = cos(theta), color = "darkred", size = 3) +
  annotate("text", x = -sin(theta) - 0.05, y = cos(theta) - 0.05, 
           label = "Sigma[4]*minute ~ group('(', list(-sqrt(2)/2, sqrt(2)/2), ')')", 
           parse = TRUE, size = 4, color = "darkred", fontface = "bold", hjust = 1) +
  
  # --- ANCILLARY ELEMENTS ---
  # CCW rotation arc and angle label
  geom_path(data = arc_data, aes(x = x, y = y), 
            arrow = arrow(length = unit(0.2, "cm"), type = "closed"), color = "black") +
  annotate("text", x = 0.55, y = 0.2, label = "theta == pi/4", parse = TRUE, size = 5) +
  
  # Plot setup
  coord_fixed(xlim = c(-1.4, 1.9), ylim = c(-0.4, 1.9)) +
  scale_fill_manual(values = c("gray30", "firebrick")) +
  scale_color_manual(values = c("gray40", "firebrick")) +
  theme_minimal() +
  theme(axis.title = element_blank(),
        axis.text = element_blank(), # Optional: hides the grid numbers if you only want arrows
        panel.grid.minor = element_blank(),
        legend.position = "none")

```

 
## Special Types of Matrices

### The Null Matrix

An $m \times n$ matrix is called a **null matrix** (or zero matrix) if all its entries are zero, and is typically denoted by $O_{m \times n}$ or simply $O$:


$$O_{m \times n} = \begin{bmatrix}
0 & 0 & \cdots & 0 \\
0 & 0 & \cdots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \cdots & 0
\end{bmatrix}$$


#### Example: A $3 \times 4$ Null Matrix

$$O_{3 \times 4} = \begin{bmatrix}
0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0
\end{bmatrix}$$


##### *Implementation in R*

A null matrix is easily generated by specifying its dimensions and filling all entries with zero.

```{r}
O <- matrix(0, nrow = 3, ncol = 4)
O
```

 

### The Square Matrix

A matrix $A$ is called a **square matrix** if $m = n$, that is, if it has an equal number of rows and columns. In this case, $A$ is said to be of **order** $n$, denoted $A_{n \times n}$ or simply $A_n$. 


$$A_{n \times n} = \begin{bmatrix}
\boldsymbol{a_{11}} & a_{12}              & \cdots & a_{1n} \\
a_{21}              & \boldsymbol{a_{22}} & \cdots & a_{2n} \\
\vdots              & \vdots              & \ddots & \vdots \\
a_{n1}              & a_{n2}              & \cdots & \boldsymbol{a_{nn}}
\end{bmatrix}$$

 

- The **main diagonal** (or principal diagonal) of this matrix consists of the entries $a_{ij}$ for which the row index equals the column index, that is, $i = j$ (shown in bold in the matrix), denoted:

$$\text{diag}(A) = \{a_{11}, a_{22}, \dots, a_{nn}\}$$

All remaining entries, where $i \neq j$, are referred to as off-diagonal elements.

::: content-box-white
**REMARK:** Although the main diagonal is most naturally associated with square matrices, it is defined for any $m \times n$ rectangular matrix, consisting of the entries $a_{ii}$ for $i = 1, 2, \ldots, \min(m, n)$.

- **Case 1: Horizontal Rectangular Matrix ($m < n$)**

For a $3 \times 4$ matrix, where $\min(3, 4) = 3$, the main diagonal contains exactly three entries. The diagonal terminates before reaching the final column.

$$A_{3 \times 4} = \begin{bmatrix}
\boldsymbol{a_{11}} & a_{12} & a_{13} & a_{14} \\
a_{21} & \boldsymbol{a_{22}} & a_{23} & a_{24} \\
a_{31} & a_{32} & \boldsymbol{a_{33}} & a_{34}
\end{bmatrix}$$

For example:

$$A_{3 \times 4} = \begin{bmatrix} \mathbf{4} & 0 & 2 & 1 \\ 3 & \mathbf{1} & 4 & 2 \\ 2 & 0 & \mathbf{1} & 3 \end{bmatrix}$$

In this instance, $\text{diag}(A) = \{4, 1, 1\}$.


- **Case 2: Vertical Rectangular Matrix ($m > n$)**

For a $4 \times 3$ matrix, where $\min(4, 3) = 3$, the diagonal also contains three entries. Here, the diagonal is exhausted before reaching the final row.

$$A_{4 \times 3} = \begin{bmatrix}
\boldsymbol{a_{11}} & a_{12} & a_{13} \\
a_{21} & \boldsymbol{a_{22}} & a_{23} \\
a_{31} & a_{32} & \boldsymbol{a_{33}} \\
a_{41} & a_{42} & a_{43}
\end{bmatrix}$$

For example:

$$A_{4 \times 3} = \begin{bmatrix} \mathbf{2} & 1 & 0 \\ 3 & \mathbf{-1} & 4 \\ 1 & 2 & \mathbf{5} \\ 0 & 3 & 1 \end{bmatrix}$$

Here, $\text{diag}(B) = \{2, -1, 5\}$. Note that the fourth row contains no diagonal entry because the index $i=4$ has no corresponding column index $j=4$.
:::

 

- In a square matrix $A$ of order $n$, the **anti-diagonal** (also known as the secondary diagonal) consists of the entries $a_{i,j}$ for which $i + j = n + 1$, running from the upper-right corner to the lower-left corner.


$$A_{n \times n} = \begin{bmatrix}
a_{11}  & a_{12} & \cdots  & a_{1 \ (n-1)} & \boldsymbol{a_{1n}} \\
a_{21}  & a_{22} & \cdots & \boldsymbol{a_{2 \ (n-1)}} & a_{2n} \\
\vdots  & \vdots & \ddots & \vdots & \vdots \\
a_{(n-1) \ 1}    & \boldsymbol{a_{(n-1) \ 2}} & \cdots & a_{(n-1) \ (n-1)} & a_{(n-1) \ n} \\
\boldsymbol{a_{n1}} & a_{n2}  & \cdots & a_{n \ (n-1)} & a_{nn}
\end{bmatrix}$$

The anti-diagonal entries are shown in bold in the matrix: 

$$\text{anti-diag}(A) = \{a_{1n}, a_{2 \ (n-1)}, \dots, a_{(n-1) \ 2}, a_{n1}\}$$

 

- The **trace** of a square matrix $A_{n \times n}$ is the sum of its **main diagonal** entries, denoted by $\text{tr}(A)$:

$$\text{tr}(A) = a_{11} + a_{22} + \cdots + a_{nn} = \sum_{i=1}^{n} a_{ii}$$

The trace is defined only for square matrices.

 

#### Example: A $3 \times 3$ Square Matrix

Consider the following square matrix of order 3:

$$A_{3 \times 3} = \begin{bmatrix} 
\mathbf{5} & 1 & 0 \\ 
3 & \mathbf{-1} & 2 \\ 
4 & 0 & \mathbf{-1}
\end{bmatrix}$$



The main diagonal entries (shown in bold) are $a_{11} = 5$, $a_{22} = -1$, and $a_{33} = -1$, hence

$$\text{diag}(A) = \{5, -1, -1\}$$  

$$\text{tr}(A) = 5 + (-1) + (-1) = 3$$


The anti-diagonal entries, for which $i + j = 4$, are $a_{13} = 0$, $a_{22} = -1$, and $a_{31} = 4$, hence

$$\text{anti-diag}(A) = \{0, -1, 4\}$$

Note that $a_{22} = -1$ lies on both the main diagonal and the anti-diagonal. This occurs in every square matrix of odd order, where the central entry belongs to both diagonals.

 

##### *Implementation in R*

We construct $A$ in R and use the built-in `diag()` function to extract its main diagonal entries. 

```{r}
entries <- c(5, 1, 0,
             3, -1, 2,
             4, 0, -1)

A <- matrix(entries, nrow = 3, ncol = 3, byrow = TRUE)
A
```




```{r}
# Main diagonal
diag(A)
```



The trace is not available as a standalone function in base R, but is straightforwardly computed as a sum.

```{r}
# Trace
sum(diag(A))
```

Extracting the anti-diagonal in R requires indexing. This can be achieved using `cbind()` to pair row and column indices appropriately:

```{r}
# Anti-diagonal 
n <- nrow(A)
A[cbind(1:n, n:1)]
```


 

### The Diagonal Matrix

A **diagonal matrix** $A$ is a square matrix in which all off-diagonal entries are zero, that is, $a_{ij} = 0$ for all $i \neq j$:

$$A_{n \times n} = \begin{bmatrix} a_{11} & 0 & \dots & 0 \\ 
0 & a_{22} & \dots & 0 \\ 
\vdots & \vdots & \ddots & \vdots \\ 
0 & 0 & \dots & a_{nn} 
\end{bmatrix}$$

 

#### Example: A $3 \times 3$ Diagonal Matrix


$$A_{3\times 3} =
  \begin{bmatrix}
    4 & 0 & 0\\
    0 & -1 & 0\\
    0 & 0 & -5
  \end{bmatrix}$$


with 

$\text{diag}(A) = \{4, -1, -5\}$

$\text{tr}(A) = 4 + (-1) + (-5) = -2$

$\text{anti-diag}(A) = \{0, -1, 0\}$

 

##### *Implementation in R*

In R, a diagonal matrix can be constructed using the `diag()` function. 


```{r}
entries <- c(4, -1, -5)

A <- diag(entries)
A
```


```{r}
# Main diagonal 
diag(A)      

```


```{r}
# Trace
sum(diag(A))
```


```{r}
# Anti-diagonal 
n <- nrow(A)
A[cbind(1:n, n:1)]
```


 

### The Anti-Diagonal Matrix

An **anti-diagonal matrix** (or skew-diagonal matrix) is a square matrix where all entries are zero except for those on the diagonal going from the lower-left corner to the upper-right corner. 

Formally, a square matrix $A$ is anti-diagonal if $a_{ij} = 0$ whenever $i + j \neq n + 1$:

$$A_{n \times n} = \begin{bmatrix}
0      & 0      & \cdots & 0      & a_{1n} \\
0      & 0      & \cdots & a_{2 \ (n-1)} & 0   \\
\vdots & \vdots & \ddots & \vdots & \vdots \\
0      & a_{(n-1) \ 2}   & \cdots & 0   & 0      \\
a_{n1} & 0      & \cdots & 0      & 0
\end{bmatrix}$$

 

::: content-box-white
**NOTE:** An anti-diagonal matrix $A$ of size $n \times n$ has nonzero entries only on the anti-diagonal (from the top-right to the bottom-left). The trace of $A$ is guaranteed to be zero if $n$ is even, but it may be non-zero if $n$ is odd.
:::


 

#### Example: A $3 \times 3$ Anti-Diagonal Matrix

In a $3 \times 3$ matrix, the anti-diagonal entries are those where $i + j = 4$:

$$A_{3\times 3} =\begin{bmatrix}
0 & 0 & 7\\
0 & 2 & 0\\
-3 & 0 & 0
\end{bmatrix}$$


with 

$\text{diag}(A) = \{0, 2, 0\}$

$\text{tr}(A) = 0 + 2 + 0 = 2$

$\text{anti-diag}(A) = \{7, 2, -3\}$

 

##### *Implementation in R*

```{r}
entries <- c(0, 0, 7,
             0, 2, 0,
            -3, 0, 0)

A <- matrix(entries, nrow = 3, byrow = TRUE)

A
```


```{r}
# Main diagonal 
diag(A)      

```


```{r}
# Trace
sum(diag(A))
```


```{r}
# Anti-diagonal 
n <- nrow(A)
A[cbind(1:n, n:1)]
```


 


### The Triangular Matrices

A square matrix $A$ is called **upper triangular** if all entries below the main diagonal are zero, that is, $a_{ij} = 0$ for all $i > j$.

$$A_{upper} = \begin{bmatrix}
a_{11} & a_{12} & \dots & a_{1n} \\
0 & a_{22} & \dots & a_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \dots & a_{nn}
\end{bmatrix}$$

Conversely, a square matrix $A$ is called **lower triangular** if all entries above the main diagonal are zero, that is, $a_{ij} = 0$ for all $i < j$.

$$A_{lower} = \begin{bmatrix}
a_{11} & 0 & \dots & 0 \\
a_{21} & a_{22} & \dots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
a_{n1} & a_{n2} & \dots & a_{nn}
\end{bmatrix}$$

 

::: content-box-white
**NOTE:** A square matrix that is simultaneously upper and lower triangular has zeros both above and below the main diagonal, and is therefore necessarily a diagonal matrix.
:::


 

#### Example: $3 \times 3$ Triangular Matrices

$$A_{upper} = \begin{bmatrix} \mathbf{2} & 3 & -1 \\ 0 & \mathbf{5} & 4 \\ 0 & 0 & \mathbf{7} \end{bmatrix}, \qquad A_{lower} = \begin{bmatrix} \mathbf{2} & 0 & 0 \\ 3 & \mathbf{5} & 0 \\ -1 & 4 & \mathbf{7} \end{bmatrix}$$
 

##### *Implementation in R*

- The upper triangular matrix

```{r}
# Define the entries for the upper matrix
entries_upper <- c(2, 3, -1, 
                   0, 5,  4, 
                   0, 0,  7)

# Create the upper matrix
A_upper <- matrix(entries_upper, nrow = 3, byrow = TRUE)
A_upper
```


- The lower triangular matrix

```{r}
# Define the entries for the lower matrix
entries_lower <- c(2, 0,  0, 
                   3, 5,  0, 
                  -1, 4,  7)


# Create the lower matrix
A_lower <- matrix(entries_lower, nrow = 3, byrow = TRUE)
A_lower
```

 

### The Identity Matrix

An **identity matrix**, typically denoted as $I$, is a square matrix whose entries are given by the Kronecker delta $\delta_{ij}$:

$$\delta_{ij} = \begin{cases} 1 & \text{if } i = j \\ 0 & \text{if } i \neq j \end{cases}$$ 

That is, it has ones on the main diagonal and zeros everywhere else. In explicit form, it is written as:

$$I_{n \times n} = \begin{bmatrix} 1 & 0 & \dots & 0 \\ 
0 & 1 & \dots & 0 \\ 
\vdots & \vdots & \ddots & \vdots \\ 
0 & 0 & \dots & 1 
\end{bmatrix}$$

Thus, $I_{n \times n}$ is a special case of a diagonal matrix where all diagonal entries equal to 1.

 

#### Example: A $3 \times 3$ Identity Matrix

The identity matrix of order 3 is written as:

$$I_{3 \times 3} =
  \begin{bmatrix}
    1 & 0 & 0\\
    0 & 1 & 0\\
    0 & 0 & 1
  \end{bmatrix}$$
  


 

##### *Implementation in R*

In R, the `diag()` function is used to generate identity matrices. When passed a single integer $n$, it returns an $n \times n$ identity matrix.

```{r}
I <- diag(3)
I
```


 

### The Permutation Matrix

A permutation matrix is a square matrix with exactly one entry equal to $1$ in each row and each column, and $0$ elsewhere. Equivalently, it is obtained by permuting the rows (or columns) of an identity matrix.

Formally, let $\sigma$ be a permutation of the set $\{1, 2, \dots, n\}$. The associated permutation matrix

$$P_{n \times n}= [p_{ij}] = \begin{bmatrix}
p_{11} & p_{12} & \dots & p_{1n} \\
p_{21} & p_{22} & \dots & p_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
p_{n1} & p_{n2} & \dots & p_{nn}
\end{bmatrix}$$

is defined entry-wise by:

$$\quad p_{ij} = \begin{cases} 1 & \text{if } j = \sigma(i) \\ 0 & \text{otherwise} \end{cases}$$

This definition states that in each row $i$, the entry $1$ is placed in column $\sigma(i)$. This is equivalent to saying that $P$ is formed by taking the rows of the identity matrix $I_n$ and reordering them according to $\sigma$.

 

#### Example: A $3 \times 3$ Permutation Matrix

Let us consider a permutation $\sigma$ of the set $\{1, 2, 3\}$ defined as:

$$\sigma(1) = 2, \quad \sigma(2) = 1, \quad \sigma(3) = 3$$

We evaluate each row ($i$) against each column ($j$):

- For row $i = 1$:

$$\begin{aligned}
j = 1 \neq \sigma(1) \rightarrow p_{11} = 0,\ \ \
j = 2 = \sigma(1) \rightarrow \boxed{p_{12} = 1}, \ \ \
j = 3 \neq \sigma(1) \rightarrow p_{13} = 0
\end{aligned}$$


- For row $i = 2$:

$$\begin{aligned}
j = 1 = \sigma(2) \rightarrow \boxed{p_{21} = 1} ,\ \ \
j = 2 \neq \sigma(2) \rightarrow p_{22} = 0 ,\ \ \
j = 3 \neq \sigma(2) \rightarrow p_{23} = 0 ,\ \ \
\end{aligned}$$


- For row $i = 3$:

$$\begin{aligned}
j = 1 \neq \sigma(3) \rightarrow p_{31} = 0,\ \ \
j = 2 \neq \sigma(3) \rightarrow p_{32} = 0,\ \ \
j = 3 = \sigma(3) \rightarrow \boxed{p_{33} = 1}
\end{aligned}$$


By arranging these values according to the given permutation, we construct the corresponding permutation matrix $P_{3 \times 3}$:

$$P = \begin{bmatrix}
0 & 1 & 0 \\ 
1 & 0 & 0 \\
0 & 0 & 1 
\end{bmatrix}$$

This matrix is obtained from the identity matrix by swapping the first and second rows, while leaving the third row unchanged. Hence, it is a permutation matrix corresponding to $\sigma$.

 

##### *Implementation in R*

```{r}
# permutation matrix (swap rows 1 and 2 of identity matrix)
P <- I[c(2, 1, 3), ]
P
```


 

### The Symmetric Matrix

A square matrix $A$ is **symmetric** if its entries satisfy:

$$a_{ij} = a_{ji} \quad \text{for all } i, j$$

This condition implies that the matrix is symmetric with respect to its main diagonal. Equivalently, interchanging its rows and columns leaves the matrix unchanged.

In its general form, the entries below the main diagonal are reflections of those above it:

$$A = \begin{bmatrix}
a_{11} & a_{12} & \dots & a_{1n} \\
a_{12} & a_{22} & \dots & a_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
a_{1n} & a_{2n} & \dots & a_{nn}
\end{bmatrix}$$


 

#### Example: A $3 \times 3$ Symmetric Matrix

Consider the following symmetric matrix $A$ of order 3:

$$A_{3\times 3} =
  \begin{bmatrix}
    13 & -4 & 2\\
    -4 & 11 & -2\\
    2 & -2 & 8
  \end{bmatrix}$$

In this matrix, the entries at positions (1,2) and (2,1) are both -4, the entries at positions (1,3) and (3,1) are both 2, and the entries at positions (2,3) and (3,2) are both -2. This reflects the symmetry property.


 

##### *Implementation in R*


```{r}
# Define entries row-wise
entries <- c(13, -4, 2,
            -4, 11, -2,
             2, -2, 8)

# Create the symmetric matrix 
A <- matrix(entries, nrow = 3)
A
```



 

### The Antisymmetric Matrix

A square matrix $A$ is **antisymmetric** (or skew-symmetric) if its entries satisfy:$$a_{ij} = -a_{ji} \quad \text{for all } i, j$$

A key property of antisymmetric matrices is that all elements on the main diagonal must be zero, because for $i=j$, the condition $a_{ii} = -a_{ii}$ can only be true if $a_{ii} = 0$.


In its general form, entries below the main diagonal are the negative reflections of those above it:

$$A = \begin{bmatrix}
0 & a_{12} & \dots & a_{1n} \\
-a_{12} & 0 & \dots & a_{2n} \\
\vdots & \vdots & \ddots & \vdots \\
-a_{1n} & -a_{2n} & \dots & 0
\end{bmatrix}$$


 

#### Example: A $3 \times 3$ Antisymmetric Matrix

Consider the following antisymmetric matrix $A$ of order 3:

$$A_{3\times 3} =\begin{bmatrix}0 & 5 & -2\\
-5 & 0 & 1\\
2 & -1 & 0
\end{bmatrix}$$

In this matrix, notice that the diagonal is all zeros. Furthermore, the entries at positions (1,2) and (2,1) are 5 and -5, (1,3) and (3,1) are -2 and 2, and (2,3) and (3,2) are 1 and -1. This reflects the antisymmetry property.


 

##### *Implementation in R*

```{r}
# Define entries row-wise
entries <- c( 0,  5, -2,
             -5,  0,  1,
              2, -1,  0)

# Create the antisymmetric matrix 
A <- matrix(entries, nrow = 3, byrow = TRUE)
A
```




## The Algebra of Matrices

### Equality of Matrices

Any meaningful algebra requires a precise notion of equality. Two matrices $A = [a_{ij}]_{m \times n}$ and $B = [b_{ij}]_{p \times q}$ are equal, written $A = B$, if and only if they have the same size and identical corresponding entries:

- $m = p$ and $n = q$
- $a_{ij} = b_{ij}$ for all $i, j$




#### Example: Equal Matrices

Let 

$$A = \begin{bmatrix} 2 & 5 \\ -1 & 0 \end{bmatrix}, \quad B = \begin{bmatrix} 2 & 5 \\ -1 & 0 \end{bmatrix}$$

Since both matrices are $2 \times 2$ and every entry matches its counterpart, we conclude that $A = B$.

 

##### *Implementation in R*

We construct both matrices and verify equality.

```{r}
A <- matrix(c(2,  5,
             -1,  0),
            nrow = 2, ncol = 2, byrow = TRUE)

B <- matrix(c(2,  5,
             -1,  0),
            nrow = 2, ncol = 2, byrow = TRUE)
```

First, we confirm that $A$ and $B$ have the same dimensions:

```{r}
# Check dimensions
dim(A) == dim(B)   # TRUE TRUE — number of rows match, number of columns match
```


Then, we verify that every corresponding entry satisfies $a_{ij} = b_{ij}$:

```{r}
# Check entry-wise equality 
all(A == B)   # TRUE — all entries match
```



