# Vector Spaces {#sec-vectors}

In physics and engineering, physical quantities are broadly classified into two fundamental types: **scalars**, which possess magnitude only (such as mass or temperature), and **vectors**, which possess both magnitude and direction (such as velocity or force). The mathematical structure that naturally describes and manipulates vectors is the concept of a **vector space**.

## Definition of Vector Space

A non-empty set $V$ is called a vector space over a field $\mathbb{K}$ if its elements are equipped with a well-defined notion of equality and two closed operations: vector addition between elements of $V$, denoted by ($+$), and scalar multiplication of a scalar in $\mathbb{K}$ with an element of $V$, denoted by ($\cdot$), satisfying the following axioms for all $\mathbf{u}, \mathbf{v}, \mathbf{w} \in V$ and all $\lambda, \mu \in \mathbb{K}$:

**Vector Addition Axioms**

1.  **Closure under Addition:** The sum of $\mathbf{u}$ and $\mathbf{v}$, denoted by $\mathbf{u} + \mathbf{v}$, is in $V$.

2.  **Additive Commutativity:** $\mathbf{u} + \mathbf{v} = \mathbf{v} + \mathbf{u}$.

3.  **Additive Associativity:** $(\mathbf{u} + \mathbf{v}) + \mathbf{w} = \mathbf{u} + (\mathbf{v} + \mathbf{w})$.

4.  **Additive Identity:** There exists a unique vector $\mathbf{0}$ in $V$, called the zero vector, such that $\mathbf{0} + \mathbf{u} =\mathbf{u} + \mathbf{0} = \mathbf{u}$ for all $u$ in $V$.

5.  **Additive Inverse:** For each $\mathbf{u}$ in $V$, there is a vector $-\mathbf{u}$ in $V$ (called the negative of $\mathbf{u}$) such that $\mathbf{u} + (-\mathbf{u}) = \mathbf{0}$.

**Scalar Multiplication Axioms**

1.  **Closure under Scalar Multiplication:** The scalar multiple of $\mathbf{u}$ by $\lambda$, denoted by $\lambda \cdot \mathbf{u}$, is in $V$.

2.  **Distributivity over Vector Addition:** $\lambda \cdot (\mathbf{u} + \mathbf{v}) = \lambda \cdot \mathbf{u} + \lambda \cdot \mathbf{v}$.

3.  **Distributivity over Scalar Addition:** $(\lambda + \mu)\cdot \mathbf{u} = \lambda\cdot \mathbf{u} + \mu\cdot \mathbf{u}$.

4.  **Compatibility of Multiplication:** $\lambda \cdot (\mu \cdot \mathbf{u}) = (\lambda \cdot \mu) \cdot \mathbf{u}$.

5.  **Scalar Identity Element:** $1 \cdot \mathbf{u} = \mathbf{u}$.

Additionally, for each $\mathbf{u}$ in $V$ and scalar $\lambda \in \mathbb{K}$, the following properties holds as an immediate consequence of the vector space axioms:

$$0 \cdot \mathbf{u} = \mathbf{0}$$ $$\lambda \cdot \mathbf{0} = \mathbf{0}$$

<br>

Unless stated otherwise, $\mathbb{K}$ is taken to be $\mathbb{R}$ or $\mathbb{C}$, and $V$ is referred to as a real or complex vector space, respectively.

::: content-box-white
**NOTE:** Axioms 1 through 5 state that $(V, +)$ forms an abelian (commutative) group. Axioms 6 through 10 define how the field $\mathbb{K}$ acts on that group.
:::

There exist many completely different types of vector spaces used throughout science and engineering:

-   **Coordinate Spaces (**$\mathbb{R}^n$ and $\mathbb{C}^n$): Ordered n-tuples of real or complex numbers, ranging from the two- and three-dimensional spaces of classical physics to the high-dimensional arrays of modern data analysis.

$$\mathbf{u} + \mathbf{v} = \begin{bmatrix} u_1 + v_1 \\ u_2 + v_2 \\ \vdots \\ u_n + v_n \end{bmatrix} \text{,} \qquad \lambda \mathbf{u} = \begin{bmatrix} \lambda u_1 \\ \lambda u_2 \\ \vdots \\ \lambda u_n \end{bmatrix}$$

-   **Matrix Spaces (**$M_{m \times n}$): The set of all $m \times n$ matrices over a field $\mathbb{K}$, where each matrix is treated as a single vector and addition and scalar multiplication are defined entry-wise,

$$\mathbf{U} + \mathbf{V} = [u_{ij} + v_{ij}] \text{,} \qquad \lambda \cdot \mathbf{U} = [\lambda \cdot u_{ij}]$$

allowing transformations, and systems of equations be represented and manipulated within a unified algebraic framework.

-   **Function Spaces (**$C[a, b]$): Continuous functions on a closed interval $[a,b]$, with addition and scaling defined pointwise,

$$(f + g)(x) = f(x) + g(x) \text{,} \qquad (\lambda \cdot f)(x) = \lambda \cdot f(x)$$

forming infinite-dimensional spaces where whole continuous functions, mathematical curves, and physical waveforms behave exactly like vectors. This forms the foundation of signal processing, acoustics, and quantum mechanics.

## Vector Subspaces

When analyzing a large vector space $V$, we are frequently interested in smaller collections of vectors within $V$ that behave as self-contained vector spaces in their own right. These collections are known as subspaces.

Formally, let $V$ be a vector space over a field $\mathbb{K}$, and let $W$ be a subset of $V$ ($W \subseteq V$). We say that $W$ is a subspace of $V$ if:

1.  It contains the zero vector: $\mathbf{0} \in W$

2.  It is closed under the inherited operation of addition: $\mathbf{u}, \mathbf{v} \in W \implies \mathbf{u} + \mathbf{v} \in W$

3.  It is closed under the inherited operation of scalar multiplication: $\mathbf{u} \in W, \lambda \in \mathbb{K} \implies \lambda\mathbf{u} \in W$

Since the remaining axioms are inherited automatically from $V$, these three conditions are both necessary and sufficient, and can be consolidated into the subspace test:

$$\mathbf{u}, \mathbf{v} \in W, \ \lambda, \mu \in \mathbb{K} \implies \lambda\mathbf{u} + \mu\mathbf{v} \in W$$

<br>

::: content-box-white
In the Cartesian space $\mathbb{R}^3$, every plane $P$ passing through the origin can be written as

$$P = \{(x,y,z) \in \mathbb{R}^3 \mid \alpha x + \beta y + \gamma z = 0\},$$

where $\alpha, \beta, \gamma \in \mathbb{R}$ are not all zero. To verify that $P$ is a subspace of $\mathbb{R}^3$, we check the three subspace properties.

**1. Contains the zero vector**

Substituting $(0,0,0)$ into the plain equation gives

$$\alpha(0) + \beta(0) + \gamma(0) = 0,$$

So the zero vector is in the plane.

**2. Closed under addition**

Let $\mathbf{u} = (x_1, y_1, z_1) \in P$ and $\mathbf{v} = (x_2, y_2, z_2) \in P$. Their sum is defined component-wise as:

$$\mathbf{u} + \mathbf{v} = (x_1 + x_2, \; y_1 + y_2, \; z_1 + z_2)$$

Since both $\mathbf{u}$ and $\mathbf{v}$ lie in the plane $P$, their coordinates must satisfy the plane's defining equation:

$$\alpha x_1 + \beta y_1 + \gamma z_1 = 0$$

and

$$\alpha x_2 + \beta y_2 + \gamma z_2 = 0$$

Adding these two equations yields:

$$\alpha(x_1 + x_2) + \beta(y_1 + y_2) + \gamma(z_1 + z_2) = 0$$

Hence,

$$(x_1 + x_2, \; y_1 + y_2, \; z_1 + z_2) \in P$$

$$\mathbf{u} + \mathbf{v} \in P$$

**3. Closed under scalar multiplication**

Let $(x_1, y_1, z_1) \in P$ and let $\lambda \in \mathbb{R}$. Since

$$\alpha x_1 + \beta y_1 + \gamma z_1 = 0,$$

multiplying by $\lambda$ gives

$$\alpha(\lambda x_1) + \beta(\lambda y_1) + \gamma(\lambda z_1) = 0$$

Therefore,$$(\lambda x_1, \lambda y_1, \lambda z_1) \in P$$

$$\lambda\mathbf{u} \in P$$

Since all three conditions are satisfied, every plane through the origin is a subspace of $\mathbb{R}^3$.

Moreover, the subspaces of $\mathbb{R}^3$ are precisely:

-   The trivial subspace $\{(0,0,0)\}$
-   All lines through the origin
-   All planes through the origin
-   $\mathbb{R}^3$ itself
:::

## Vector Space in $\mathbb{R}^n$

We can represent vectors as a sequence of entries (or elements) arranged in a column, called a **column vector**. For example, we write a vector $\mathbf{u}$ in $\mathbb{R}^n$ as

$$\mathbf{u} = \begin{bmatrix} u_1 \\ u_2 \\ \vdots \\ u_n \end{bmatrix}$$

Because writing vertically takes up a lot of space, this exact same vector is often written horizontally using parentheses and commas:

$$\mathbf{u} = (u_1, u_2, \dots, u_n)$$

 

#### Example: Vector $u = (u_1, u_2)$

For example, a vector $\mathbf{u}$ in a two-dimensional space ($\mathbb{R}^2$) can be expressed algebraically as an ordered pair of its entries, $u_1 = 1.5$ and $u_2 = 1$, written as $\mathbf{u} = (1.5, 1)$. Geometrically, it is represented as an arrow, or directed line segment, starting at the origin in the coordinate plane and ending at the point with coordinates $u_1$ and $u_2$ (@fig-vector).

```{r}
#| echo: false
#| warning: false
#| fig-align: center
#| label: fig-vector
#| fig-cap: 'In a Cartesian coordinate system, a vector in $\mathbb{R}^2$ is geometrically represented as a directed line segment (an arrow).'


library(ggplot2)

# --- 1. Define Vector and Label Data ---
u1 <- 1.5
u2 <- 1.0

point_data <- data.frame(
  x = u1,
  y = u2,
  label = "u == (list(1.5, 1))"
)

# --- 2. Build the Plot ---
ggplot() +
  # Clear theme to remove background grid lines completely
  theme_void() + 
  theme(
    plot.margin = margin(20, 20, 20, 20)
  ) +
  
  # 1. Vertical and Horizontal Dashed Projection Lines (Palette from Image 1)
  geom_segment(aes(x = u1, y = u2, xend = u1, yend = 0), 
               color = "#555555", lty = "dashed", size = 0.5) +
  geom_segment(aes(x = u1, y = u2, xend = 0, yend = u2), 
               color = "#555555", lty = "dashed", size = 0.5) +

  
  # 2. Axis Lines (x and y) using the gray color (#555555)
  geom_segment(aes(x = 0, y = 0, xend = 2.0, yend = 0), 
               arrow = arrow(length = unit(0.25, "cm"), type = "open"), 
               color = "#555555", size = 0.5) +
  geom_segment(aes(x = 0, y = 0, xend = 0, yend = 1.8), 
               arrow = arrow(length = unit(0.25, "cm"), type = "open"), 
               color = "#555555", size = 0.5) +
  
  # Axis Labels (Muted gray text)
  annotate("text", x = 2.15, y = 0, label = "x", fontface = "italic", color = "#555555", size = 6) +
  annotate("text", x = 0, y = 1.95, label = "y", fontface = "italic", color = "#555555", size = 6) +
  annotate("text", x = -0.05, y = -0.08, label = "O (0,0)", fontface = "italic", color = "#555555", size = 5) +
  
  # Axis Ticks / Intercept Labels for the components
  annotate("text", x = u1, y = -0.1, label = "u[1] == 1.5", parse = TRUE, color = "#555555", size = 5) +
  annotate("text", x = -0.17, y = u2, label = "u[2] == 1", parse = TRUE, color = "#555555", size = 5) +
  
  # 3. The Vector Arrow using the Dark Red outline color (#8B0000)
  geom_segment(aes(x = 0, y = 0, xend = u1, yend = u2), 
               arrow = arrow(length = unit(0.35, "cm"), type = "closed"), 
               color = "#8B0000", size = 1.0) +
  
  # Vector Name Label
  annotate("text", x = u1 * 0.45, y = u2 * 0.55 + 0.1, label = "u", fontface = "bold", color = "#8B0000", size = 5) +
  
  
  # 5. Math Coordinate Label matching the dark red style
  geom_text(data = point_data, aes(x = x, y = y, label = label), 
            color = "#8B0000", parse = TRUE, size = 5, hjust = -0.15, vjust = -0.15) +
  
  # 6. Fix coordinate ratio and set plot boundaries
  coord_fixed(xlim = c(-0.3, 2.5), ylim = c(-0.3, 2.1))
```

##### *Implementation in R*

We can initialize our specific vector $\mathbf{u} = (1.5, 1)$ and print its entries to the console as follows:

```{r}
u <- c(1.5, 1)
u
```

## Vector Addition in $\mathbb{R}^n$

If $\mathbf{a} = (a_1, a_2, \dots, a_n)$ and $\mathbf{b} = (b_1, b_2, \dots, b_n)$ are two vectors in $\mathbb{R}^n$, then the sum of $\mathbf{a}$ and $\mathbf{b}$ is the vector:$$\mathbf{a} + \mathbf{b} = (a_1 + b_1, a_2 + b_2, \dots, a_n + b_n)$$

We add two vectors, $\mathbf{u} = (u_1, u_2)$ and $\mathbf{v} = (v_1, v_2)$, by adding their components:

$$u + v = (u_1, u_2) + (v_1, v_2) = (u_1 + v_1, u_2 + v_2)$$

```{r}
#| echo: false
#| warning: false
#| fig-align: center
#| label: fig-vector2
#| fig-cap: 'In a Cartesian coordinate system, a vector in $\mathbb{R}^2$ is geometrically represented as a directed line segment (an arrow).'


# 1. Define vector components
ux <- 3; uy <- 1
vx <- 1; vy <- 2
rx <- ux + vx  # 4
ry <- uy + vy  # 3

# 2. Build the blank theme canvas
ggplot() +
  # Custom Axis Lines (Charcoal Gray)
  geom_segment(aes(x = -0.5, y = 0, xend = 4.8, yend = 0), 
               arrow = arrow(length = unit(0.2, "cm"), type = "open"), color = "#555555", linewidth = 0.6) +
  geom_segment(aes(x = 0, y = -0.5, xend = 0, yend = 3.8), 
               arrow = arrow(length = unit(0.2, "cm"), type = "open"), color = "#555555", linewidth = 0.6) +
  
  # Axis Labels
  annotate("text", x = 4.7, y = -0.2, label = "x", fontface = "italic", size = 4.5, color = "#555555") +
  annotate("text", x = -0.2, y = 3.7, label = "y", fontface = "italic", size = 4.5, color = "#555555") +
  annotate("text", x = -0.15, y = -0.15, label = "O", fontface = "italic", size = 4, color = "#555555") +
  
  # Parallelogram / Projection Guidelines (Dashed Lines)
  geom_segment(aes(x = ux, y = uy, xend = rx, yend = ry), lty = "dashed", color = "#777777", linewidth = 0.5) +
  geom_segment(aes(x = vx, y = vy, xend = rx, yend = ry), lty = "dashed", color = "#777777", linewidth = 0.5) +
  
  # Vector 'u' (Blue)
  geom_segment(aes(x = 0, y = 0, xend = ux, yend = uy), 
               arrow = arrow(length = unit(0.25, "cm"), type = "closed"), color = "#1f77b4", linewidth = 1.0) +
  # Vector 'v' (Green)
  geom_segment(aes(x = 0, y = 0, xend = vx, yend = vy), 
               arrow = arrow(length = unit(0.25, "cm"), type = "closed"), color = "#2ca02c", linewidth = 1.0) +
  # Resultant Vector 'u + v' (Maroon/Red)
  geom_segment(aes(x = 0, y = 0, xend = rx, yend = ry), 
               arrow = arrow(length = unit(0.3, "cm"), type = "closed"), color = "#8B0000", linewidth = 1.2) +
  
  # Mathematical Labels (Parsed textmath expressions)
  annotate("text", x = 1.7, y = 0.3, label = "u", fontface = "bold.italic", size = 5, color = "#1f77b4") +
  annotate("text", x = 0.3, y = 1.2, label = "v", fontface = "bold.italic", size = 5, color = "#2ca02c") +
  annotate("text", x = 1.8, y = 1.8, label = "u + v", fontface = "bold.italic", size = 5, color = "#8B0000") +
  
  # Terminal Coordinate Labels
  annotate("text", x = ux + 0.4, y = uy, label = "u == (list(3, 1))", parse = TRUE, size = 4, color = "#1f77b4") +
  annotate("text", x = vx - 0.4, y = vy + 0.1, label = "v == (list(1, 2))", parse = TRUE, size = 4, color = "#2ca02c") +
  annotate("text", x = rx + 0.7, y = ry, label = "w == (list(4, 3))", parse = TRUE, size = 4.2, color = "#8B0000") +
  
  # Force layout constraints matching original graphic style
  coord_fixed() +
  theme_void()
```

 

#### Example: Vector addition u + v

$$w = u + v = (3, 1) + (1, 2) = (3 + 1, 1 + 2) = (4, 3)$$

##### *Implementation in R*

```{r}
u <- c(3, 1)
v <- c(1, 2)

w <- u + v
w
```
