# Vectors {#sec-vectors}

We can represent vectors as a sequence of entries (or elements) arranged in a column, called a **column vector**. For example, we write a vector $\mathbf{u}$ in $\mathbb{R^n}$ as

$$\mathbf{u} = \begin{bmatrix} u_1 \\ u_2 \\ \vdots \\ u_n \end{bmatrix}$$

Because writing vertically takes up a lot of space, this exact same vector is often written horizontally using parentheses and commas:

$$\mathbf{u} = (u_1, u_2, \dots, u_n)$$


 

#### Example: Matrix Product

For example, a vector $\mathbf{u}$ in a two-dimensional space ($\mathbb{R}^2$) can be expressed algebraically as an ordered pair of its entries, $u_1 = 1.5$ and $u_2 = 1$, written as $\mathbf{u} = (1.5, 1)$. Geometrically, it is represented as a directed arrow spanning from the origin within a Cartesian coordinate system (@fig-vector).


```{r}
#| echo: false
#| warning: false
#| fig-align: center
#| label: fig-vector
#| fig-cap: 'In a Cartesian coordinate system, a vector in $\mathbb{R}^2$ is geometrically represented as a directed line segment (an arrow).'


library(ggplot2)

# --- 1. Define Vector and Label Data ---
u1 <- 1.5
u2 <- 1.0

point_data <- data.frame(
  x = u1,
  y = u2,
  label = "u == (list(1.5, 1))"
)

# --- 2. Build the Plot ---
ggplot() +
  # Clear theme to remove background grid lines completely
  theme_void() + 
  theme(
    plot.margin = margin(20, 20, 20, 20)
  ) +
  
  # 1. Vertical and Horizontal Dashed Projection Lines (Palette from Image 1)
  geom_segment(aes(x = u1, y = u2, xend = u1, yend = 0), 
               color = "#555555", lty = "dashed", size = 0.5) +
  geom_segment(aes(x = u1, y = u2, xend = 0, yend = u2), 
               color = "#555555", lty = "dashed", size = 0.5) +

  
  # 2. Axis Lines (x and y) using the gray color (#555555)
  geom_segment(aes(x = 0, y = 0, xend = 2.0, yend = 0), 
               arrow = arrow(length = unit(0.25, "cm"), type = "open"), 
               color = "#555555", size = 0.5) +
  geom_segment(aes(x = 0, y = 0, xend = 0, yend = 1.8), 
               arrow = arrow(length = unit(0.25, "cm"), type = "open"), 
               color = "#555555", size = 0.5) +
  
  # Axis Labels (Muted gray text)
  annotate("text", x = 2.15, y = 0, label = "x", fontface = "italic", color = "#555555", size = 6) +
  annotate("text", x = 0, y = 1.95, label = "y", fontface = "italic", color = "#555555", size = 6) +
  annotate("text", x = -0.05, y = -0.08, label = "O (0,0)", fontface = "italic", color = "#555555", size = 5) +
  
  # Axis Ticks / Intercept Labels for the components
  annotate("text", x = u1, y = -0.1, label = "u[1] == 1.5", parse = TRUE, color = "#555555", size = 5) +
  annotate("text", x = -0.17, y = u2, label = "u[2] == 1", parse = TRUE, color = "#555555", size = 5) +
  
  # 3. The Vector Arrow using the Dark Red outline color (#8B0000)
  geom_segment(aes(x = 0, y = 0, xend = u1, yend = u2), 
               arrow = arrow(length = unit(0.35, "cm"), type = "closed"), 
               color = "#8B0000", size = 1.0) +
  
  # Vector Name Label
  annotate("text", x = u1 * 0.45, y = u2 * 0.55 + 0.1, label = "u", fontface = "bold", color = "#8B0000", size = 5) +
  
  
  # 5. Math Coordinate Label matching the dark red style
  geom_text(data = point_data, aes(x = x, y = y, label = label), 
            color = "#8B0000", parse = TRUE, size = 5, hjust = -0.15, vjust = -0.15) +
  
  # 6. Fix coordinate ratio and set plot boundaries
  coord_fixed(xlim = c(-0.3, 2.5), ylim = c(-0.3, 2.1))
```



##### *Implementation in R*

We can initialize our specific vector $\mathbf{u} = (1.5, 1)$ and print its entries to the console as follows:


```{r}
u <- c(1.5, 1)
u
```








