# Vector Spaces {#sec-vectors}

In physical sciences, measurable quantities are generally classified into two fundamental types: scalars and vectors. A **scalar** is completely specified by its magnitude alone, that is, by a single numerical value, as in the case of mass or temperature. In contrast, a **vector** is characterized by both magnitude and direction, as in velocity and force. The mathematical framework that provides a rigorous description of vectors and the operations defined on them is the concept of a **vector space**, a fundamental structure in linear algebra.

## Vectors in the Plane

A vector in the plane is represented geometrically by an arrow, i.e., a directed line segment whose initial point is the origin and whose terminal point is a point in the plane. The direction of the arrow indicates the direction of the vector and the length of the arrow represents its magnitude (@fig-vector00).

In this book, vectors are represented by lowercase letters set in boldface type, such as $\mathbf{u}$. However, when we wish to specify that a vector $\mathbf{u}$ has initial point A and terminal point B, we denote it by $\mathbf{u} = \overrightarrow{AB}$.

```{r}
#| echo: false
#| warning: false
#| fig-asp: 0.6
#| fig-width: 6
#| fig-height: 4
#| fig-align: center
#| label: fig-vector00
#| fig-cap: 'In a Cartesian coordinate system, a vector in $\mathbb{R}^2$ is geometrically represented as a directed line segment (an arrow).'


library(ggplot2)

# --- 1. Define Vector Component and Label Data ---
u1 <- 1.5
u2 <- 1.0

# Data frame for the terminal point math label
point_data <- data.frame(
  x = u1,
  y = u2,
  label = "(list(u[1], u[2]))"
)

# --- 2. Build the Plot ---
ggplot() +
  # Clear theme to remove background grid lines completely
  theme_void() + 
  theme(
    plot.margin = margin(20, 20, 20, 20)
  ) +
  
  # 1. Vertical and Horizontal Dashed Projection Lines
  #geom_segment(aes(x = u1, y = u2, xend = u1, yend = 0), 
              # color = "#777777", lty = "dashed", linewidth = 0.5) +
  #geom_segment(aes(x = u1, y = u2, xend = 0, yend = u2), 
               #color = "#777777", lty = "dashed", linewidth = 0.5) +
  
  # 2. Axis Lines (x and y) with open arrow heads
  #geom_segment(aes(x = -0.2, y = 0, xend = 2.2, yend = 0), 
              # arrow = arrow(length = unit(0.25, "cm"), type = "open"), 
               #color = "#555555", linewidth = 0.6) +
  #geom_segment(aes(x = 0, y = -0.2, xend = 0, yend = 1.8), 
               #arrow = arrow(length = unit(0.25, "cm"), type = "open"), 
               #color = "#555555", linewidth = 0.6) +
  
  # Axis Labels (Muted dark gray text in italics)
  #annotate("text", x = 2.3, y = 0, label = "x", fontface = "italic", color = "#333333", size = 6) +
  #annotate("text", x = 0, y = 1.95, label = "y", fontface = "italic", color = "#333333", size = 6) +
  
  # Axis Ticks / Intercept Labels for components
  #annotate("text", x = u1, y = -0.12, label = "u[1]", parse = TRUE, color = "#555555", size = 5) +
  #annotate("text", x = -0.12, y = u2, label = "u[2]", parse = TRUE, color = "#555555", size = 5) +
  
  # 3. The Vector Arrow (Premium Dark Red, thick line, closed arrow head)
  geom_segment(aes(x = 0, y = 0, xend = u1 - 0.001, yend = u2 - 0.001), 
               arrow = arrow(length = unit(0.35, "cm"), type = "closed"), 
               color = "#1f77b4", linewidth = 1.2) +
  
  # Vector Name Label (Boldface type)
  annotate("text", x = u1 * 0.4, y = u2 * 0.5 + 0.08, label = "bold(u)", parse = TRUE, color = "#1f77b4", size = 6) +
  
  # 4. Initial & Terminal Geometric Points
  geom_point(aes(x = 0, y = 0), color = "black", size = 2.5) +    # Initial Point Dot
  geom_point(aes(x = u1, y = u2), color = "black", size = 2.5) + # Terminal Point Dot
  
  # Point Structural Text Labels
  annotate("text", x = -0.08, y = -0.08, label = "A", color = "black", size = 5) +
  annotate("text", x = -0.25, y = -0.22, label = "Initial Point (Origin)", color = "black", size = 3.5, hjust = 0) +
  annotate("text", x = u1 + 0.05, y = u2 - 0.1, label = "Terminal Point", color = "black", size = 3.5, hjust = 0) +
  annotate("text", x = u1 + 0.08, y = u2 + 0.08, label = "B", color = "black", size = 5) +

  
  # Math Coordinate Coordinate Label matching the dark red vector style
  #geom_text(data = point_data, aes(x = x, y = y, label = label), 
            #color = "#8B0000", parse = TRUE, size = 5, hjust = -0.05, vjust = -0.6) +
  
  # 5. Fix coordinate ratio and set clean boundaries
  coord_fixed(xlim = c(-0.4, 2.0), ylim = c(-0.4, 1.5)) +
  theme(
  plot.margin = margin(t = 0, r = 0, b = 0, l = 0)
)
```

<br>

From elementary physics, we know that two forces, $\mathbf{F}_1$ and $\mathbf{F}_2$, acting on the same object and separated by an angle $\phi$, can be combined into a single resultant force given by the vector sum $\mathbf{F}_1 + \mathbf{F}_2$. According to the parallelogram law of forces, this resultant is represented geometrically by the directed diagonal of the parallelogram constructed with $\mathbf{F}_1$ and $\mathbf{F}_2$ as adjacent sides (@fig-vector01 **a**).

We also know that if we multiply force $\mathbf{F}_1$ by a scalar $\lambda$, we obtain a force $\lambda \cdot \mathbf{F}_1$ whose magnitude is scaled by $|\lambda|$. If $\lambda > 0$, the direction is preserved; if $\lambda < 0$, the direction is reversed; and if $\lambda = 0$, the result is the zero vector $\mathbf{0}$ (@fig-vector01 **b**).

<br>

```{r}
#| echo: false
#| warning: false
#| fig-asp: 0.6
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
#| label: fig-vector01
#| fig-cap: '$\textbf{(a) Vector addition:}$ the blue vector, $\mathbf{F_{1}} + \mathbf{F_{2}}$, is the diagonal of the parallelogram spanned by $\mathbf{F_{1}}$ and $\mathbf{F_{2}}$, where $\varphi$ denotes the angle between them. $\textbf{(b) Scalar multiplication:}$ $\lambda \cdot \mathbf{F_{1}}$ is parallel to $\mathbf{F_{1}}$ and scaled by a factor of $|\lambda|$. The vector maintains its original direction if $\lambda > 0$ and is completely reversed if $\lambda < 0$.'

library(ggplot2)
library(patchwork)

# ── Aesthetic Configurations ──────────────────────────────────────────────────
col_F1  <- "#534AB7"   # Purple
col_F2  <- "#0F6E56"   # Teal
col_R   <- "#993C1D"   # Coral
col_arc <- "grey50"
font_layout <- "serif" # Uniform mathematical typography

# ── Shared Vector Data ────────────────────────────────────────────────────────
F1  <- c(3, 4)
F2  <- c(4, -1)
R   <- F1 + F2

# Scalar factors for Panel 2
λ_pos <- 1.4
λ_neg <- -0.8

F1_pos <- λ_pos * F1
F1_neg <- λ_neg * F1

# Coordinate Envelopes
xlim_p1 <- c(-0.5, 8.0);  ylim_p1 <- c(-1.5, 5.5)
xlim_p2 <- c(-0.5, 10.0); ylim_p2 <- c(-0.6, 7.2) # Balanced vertical envelope

# ── PANEL 1: VECTOR ADDITION ─────────────────────────────────────────────────
add_df <- data.frame(
  x = c(0, 0, 0), y = c(0, 0, 0),
  xend = c(F1[1], F2[1], R[1]), yend = c(F1[2], F2[2], R[2]),
  col  = c(col_F1, col_F2, col_R)
)

ghost_df <- data.frame(
  x = c(F1[1], F2[1]), y = c(F1[2], F2[2]),
  xend = c(R[1], R[1]), yend = c(R[2], R[2])
)

angle_F1  <- atan2(F1[2], F1[1])
angle_F2  <- atan2(F2[2], F2[1])
arc_r     <- 0.85
arc_df    <- data.frame(
  x = arc_r * cos(seq(angle_F2, angle_F1, length.out = 180)),
  y = arc_r * sin(seq(angle_F2, angle_F1, length.out = 180))
)
mid_angle <- (angle_F1 + angle_F2) / 2

p1 <- ggplot() +
  geom_segment(
    data = ghost_df, aes(x = x, y = y, xend = xend, yend = yend),
    linetype = "22", colour = "grey75", linewidth = 0.55
  ) +
  geom_path(data = arc_df, aes(x = x, y = y), colour = col_arc, linewidth = 0.45) +
  annotate(
    "text", x = 1.20 * cos(mid_angle), y = 0.95 * sin(mid_angle),
    label = "phi", parse = TRUE, size = 4.5, colour = col_arc, family = font_layout
  ) +
  geom_segment(
    data = add_df, aes(x = x, y = y, xend = xend, yend = yend),
    colour = add_df$col, arrow = arrow(length = unit(0.22, "cm"), type = "closed"),
    linewidth = 1.3, linejoin = "mitre"
  ) +
  annotate("text", x = F1[1]/2 - 0.45, y = F1[2]/2 + 0.35, label = "bold(F)[1]", parse = TRUE, size = 4.8, colour = col_F1, family = font_layout) +
  annotate("text", x = F2[1]/2 + 0.45, y = F2[2]/2 + 0.25, label = "bold(F)[2]", parse = TRUE, size = 4.8, colour = col_F2, family = font_layout) +
  annotate("text", x = R[1]/2 + 0.85,  y = R[2]/2 - 0.35, label = "bold(F)[1] + bold(F)[2]", parse = TRUE, size = 4.8, colour = col_R, family = font_layout) +
  coord_equal(xlim = xlim_p1, ylim = ylim_p1, expand = FALSE) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12, margin = margin(b = 10, t = 10))
  ) +
  labs(title = "(a) Vector Addition")


# ── PANEL 2: SCALAR MULTIPLICATION (MIXED APEX/BOTTOM LABELS) ─────────────────
ox1 <- 0     # Origin X for F1
ox2 <- 2.2   # Origin X for λ > 0
ox3 <- 4.7   # Origin X for λ < 0 (Reversed origin)

scl_df <- data.frame(
  x    = c(ox1, ox2, ox3 + F1_pos[1]), 
  y    = c(0,   0,   F1_pos[2]),       
  xend = c(ox1 + F1[1], ox2 + F1_pos[1], ox3), 
  yend = c(F1[2],       F1_pos[2],       0)    
)

p2 <- ggplot() +
  # Zone bounding visual reference margins
  annotate("segment", x = ox1 - 0.5, y = 0, xend = ox3 + 4.0, yend = 0, 
           colour = "grey85", linetype = "dotted", linewidth = 0.5) +
  annotate("segment", x = ox2 - 0.5, y = F1_pos[2], xend = ox3 + 4.0, yend = F1_pos[2], 
           colour = "grey85", linetype = "dotted", linewidth = 0.5) +
  # Vector Parallel Series
  geom_segment(
    data = scl_df, aes(x = x, y = y, xend = xend, yend = yend),
    colour = col_F1, arrow = arrow(length = unit(0.22, "cm"), type = "closed"),
    linewidth = 1.3, linejoin = "mitre"
  ) +
  # High-efficiency labels located cleanly at respective tops/bottoms
  # F1 label (Top)
  annotate("text", x = ox1 + F1[1] - 0.10, y = F1[2] + 0.40, 
           label = "bold(F)[1]", parse = TRUE, size = 4.6, colour = col_F1, family = font_layout) +
  # λ > 0 label (Top peak)
  annotate("text", x = ox2 + F1_pos[1], y = F1_pos[2] + 0.40, 
           label = "lambda * bold(F)[1] ~~~ (lambda > 0)", parse = TRUE, size = 4.6, colour = col_F1, family = font_layout) +
  # λ < 0 label (Placed at the bottom baseline where the arrow head points)
  annotate("text", x = ox3 + 1.20, y = -0.40, 
           label = "lambda * bold(F)[1] ~~~ (lambda < 0)", parse = TRUE, size = 4.6, colour = col_F1, family = font_layout) +
  # Constraints & Minimal Theme Execution
  coord_equal(xlim = xlim_p2, ylim = ylim_p2, expand = FALSE) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12, margin = margin(b = 10, t = 10))
  ) +
  labs(title = "(b) Scalar Multiplication")


# ── Combined Composition Layout ───────────────────────────────────────────────
p1 + plot_spacer() + p2 + 
  plot_layout(widths = c(1, 0.05, 1.2)) & 
  theme(plot.background = element_rect(fill = "white", colour = NA))
```

Vector addition and scalar multiplication, together with a specific set of properties, define the foundational framework of a vector space.

## Definition of Vector Space

A non-empty set $V$ is called a vector space over a field $\mathbb{K}$ if its elements are equipped with a well-defined notion of equality, denoted by ($=$), and two closed operations: vector addition between elements of $V$, denoted by ($+$), and scalar multiplication of a scalar in $\mathbb{K}$ with an element of $V$, denoted by ($\cdot$), satisfying the following axioms for all $\mathbf{u}, \mathbf{v}, \mathbf{w} \in V$ and all $\lambda, \mu \in \mathbb{K}$:

**Vector Addition Axioms**

1.  **Closure under Addition:** The sum of $\mathbf{u}$ and $\mathbf{v}$, denoted by $\mathbf{u} + \mathbf{v}$, is in $V$.

2.  **Additive Commutativity:** $\mathbf{u} + \mathbf{v} = \mathbf{v} + \mathbf{u}$.

3.  **Additive Associativity:** $(\mathbf{u} + \mathbf{v}) + \mathbf{w} = \mathbf{u} + (\mathbf{v} + \mathbf{w})$.

4.  **Additive Identity:** There exists a unique vector $\mathbf{0}$ in $V$, called the zero vector, such that $\mathbf{0} + \mathbf{u} =\mathbf{u} + \mathbf{0} = \mathbf{u}$ for all $u$ in $V$.

5.  **Additive Inverse:** For each $\mathbf{u}$ in $V$, there is a vector $-\mathbf{u}$ in $V$ (called the opposite of $\mathbf{u}$) such that $\mathbf{u} + (-\mathbf{u}) = \mathbf{0}$.

**Scalar Multiplication Axioms**

6.  **Closure under Scalar Multiplication:** The scalar multiple of $\mathbf{u}$ by $\lambda$, denoted by $\lambda \cdot \mathbf{u}$, is in $V$.

7.  **Distributivity over Vector Addition:** $\lambda \cdot (\mathbf{u} + \mathbf{v}) = \lambda \cdot \mathbf{u} + \lambda \cdot \mathbf{v}$.

8.  **Distributivity over Scalar Addition:** $(\lambda + \mu)\cdot \mathbf{u} = \lambda\cdot \mathbf{u} + \mu\cdot \mathbf{u}$.

9.  **Compatibility of Multiplication:** $\lambda \cdot (\mu \cdot \mathbf{u}) = (\lambda \cdot \mu) \cdot \mathbf{u}$.

10. **Scalar Identity Element:** $1 \cdot \mathbf{u} = \mathbf{u}$.

Additionally, for each $\mathbf{u}$ in $V$ and scalar $\lambda \in \mathbb{K}$, the following properties holds as an immediate consequence of the vector space axioms:

$$0 \cdot \mathbf{u} = \mathbf{0}$$ $$\lambda \cdot \mathbf{0} = \mathbf{0}$$

<br>

Unless stated otherwise, $\mathbb{K}$ is taken to be $\mathbb{R}$ or $\mathbb{C}$, and $V$ is referred to as a real or complex vector space, respectively.

::: content-box-white
**NOTE:** Axioms 1 through 5 state that $(V, +)$ forms an abelian (commutative) group. Axioms 6 through 10 define how the field $\mathbb{K}$ acts on that group.
:::

There exist many completely different types of vector spaces used throughout science and engineering, including:

-   **Coordinate Spaces** ($\mathbb{R}^n$ and $\mathbb{C}^n$): Ordered n-tuples of real or complex numbers, ranging from the two- and three-dimensional spaces of classical physics to the high-dimensional arrays of modern data analysis.

$$\mathbf{u} + \mathbf{v} = \begin{bmatrix} u_1 + v_1 \\ u_2 + v_2 \\ \vdots \\ u_n + v_n \end{bmatrix} \text{,} \qquad \lambda \cdot \mathbf{u} = \begin{bmatrix} \lambda \cdot u_1 \\ \lambda \cdot u_2 \\ \vdots \\ \lambda \cdot u_n \end{bmatrix}$$

-   **Matrix Spaces** ($M_{m \times n}$): The set of all $m \times n$ matrices over a field $\mathbb{K}$, where each matrix is treated as a single vector and addition and scalar multiplication are defined entry-wise,

$$\mathbf{U} + \mathbf{V} = [u_{ij} + v_{ij}] \text{,} \qquad \lambda \cdot \mathbf{U} = [\lambda \cdot u_{ij}]$$

allowing transformations, and systems of equations be represented and manipulated within a unified algebraic framework.

-   **The Vector Space of Polynomials** ($\mathbb{P}_n$): the set of all polynomials of degree at most $n$ with coefficients from a field $\mathbb{K}$ (typically $\mathbb{R}$ or $\mathbb{C}$), along with the zero polynomial. Vectors in this space are algebraic expressions, where addition and scalar multiplication are performed by combining coefficients of like powers of $x$,

$$(p + q)(x) = \sum_{i=0}^{n} (p_i + q_i)x^i \text{,} \qquad (\lambda \cdot p)(x) = \sum_{i=0}^{n} (\lambda \cdot p_i)x^i$$

-   **Function Spaces** ($C[a, b]$): Continuous functions on a closed interval $[a,b]$, with addition and scaling defined pointwise,

$$(f + g)(x) = f(x) + g(x) \text{,} \qquad (\lambda \cdot f)(x) = \lambda \cdot (f(x))$$

forming infinite-dimensional spaces where whole continuous functions, mathematical curves, and physical waveforms behave exactly like vectors. This forms the foundation of signal processing, acoustics, and quantum mechanics.

## Vector Subspaces

When analyzing a large vector space $V$, we are frequently interested in smaller collections of vectors within $V$ that behave as self-contained vector spaces in their own right. These collections are known as subspaces.

Formally, let $V$ be a vector space over a field $\mathbb{K}$, and let $W$ be a subset of $V$ ($W \subseteq V$). We say that $W$ is a subspace of $V$ if:

1.  It contains the zero vector: $\mathbf{0} \in W$

2.  It is closed under the inherited operation of addition: $\mathbf{u}, \mathbf{v} \in W \implies \mathbf{u} + \mathbf{v} \in W$

3.  It is closed under the inherited operation of scalar multiplication: $\mathbf{u} \in W, \lambda \in \mathbb{K} \implies \lambda\mathbf{u} \in W$

Since the remaining axioms are inherited automatically from $V$, these three conditions are both necessary and sufficient, and can be consolidated into the subspace test:

$$\mathbf{u}, \mathbf{v} \in W, \ \lambda, \mu \in \mathbb{K} \implies \lambda\mathbf{u} + \mu\mathbf{v} \in W$$

<br>

::: content-box-white
**Example**

In the Cartesian space $\mathbb{R}^3$, every plane $P$ passing through the origin can be written as

$$P = \{(x,y,z) \in \mathbb{R}^3 \mid \alpha x + \beta y + \gamma z = 0\},$$

where $\alpha, \beta, \gamma \in \mathbb{R}$ are not all zero. To verify that $P$ is a subspace of $\mathbb{R}^3$, we check the three subspace properties.

```{r}
#| echo: false
#| warning: false
#| fig-align: center
#| out-width: "100%"

library(plotly)
library(dplyr) 

# ============================================================
#  PARAMETERS & CONFIGURATION (Subspace Plane passing through 0)
# ============================================================
alpha <- 1
beta  <- 1
gamma <- 1
r     <- 30  

# ============================================================
#  PLANE BASIS VECTORS (spanning the null space of n)
# ============================================================
find_basis <- function(a, b, g) {
  n <- c(a, b, g)
  if (abs(a) > 1e-10 || abs(b) > 1e-10) {
    e1 <- c(-b, a, 0)
  } else {
    e1 <- c(1, 0, 0)
  }
  e1 <- e1 / sqrt(sum(e1^2))
  
  # Cross product: n x e1
  e2 <- c(n[2]*e1[3]-n[3]*e1[2],
          n[3]*e1[1]-n[1]*e1[3],
          n[1]*e1[2]-n[2]*e1[1])
  e2 <- e2 / sqrt(sum(e2^2))
  list(e1 = e1, e2 = e2)
}

basis <- find_basis(alpha, beta, gamma)
e1    <- basis$e1
e2    <- basis$e2

# ============================================================
#  PLANE MESH (Continuous Surface Parametric Generation)
# ============================================================
s_vals <- seq(-r, r, length.out = 45)
t_vals <- seq(-r, r, length.out = 45)

Xm <- outer(s_vals, t_vals, function(s, t) s*e1[1] + t*e2[1])
Ym <- outer(s_vals, t_vals, function(s, t) s*e1[2] + t*e2[2])
Zm <- outer(s_vals, t_vals, function(s, t) s*e1[3] + t*e2[3])

# ============================================================
#  BACKGROUND COORD AXES & TICK MARKS
# ============================================================
axes_lines <- list(
  x = c(-25, 25, NA, 0, 0, NA, 0, 0),
  y = c(0, 0, NA, -15, 15, NA, 0, 0),
  z = c(0, 0, NA, 0, 0, NA, -12, 12)
)

off  <- 0.4
tk_x <- c(); tk_y <- c(); tk_z <- c()

for (v in c(-20, -10, 10, 20)) {
  tk_x <- c(tk_x, v, v, NA); tk_y <- c(tk_y, -off, off, NA); tk_z <- c(tk_z, 0, 0, NA)
}
for (v in c(-10, -5, 5, 10)) {
  tk_x <- c(tk_x, -off, off, NA); tk_y <- c(tk_y, v, v, NA); tk_z <- c(tk_z, 0, 0, NA)
}
for (v in c(-10, -8, -6, -4, -2, 2, 4, 6, 8, 10)) {
  tk_x <- c(tk_x, -off, off, NA); tk_y <- c(tk_y, 0, 0, NA); tk_z <- c(tk_z, v, v, NA)
}

# ============================================================
#  PLANE STRUCTURAL MESH WIREFRAME GENERATION 
# ============================================================
grid_x <- c(); grid_y <- c(); grid_z <- c()
g_lines <- seq(-r, r, by = 3)

for (t in g_lines) {
  grid_x <- c(grid_x, s_vals * e1[1] + t * e2[1], NA)
  grid_y <- c(grid_y, s_vals * e1[2] + t * e2[2], NA)
  grid_z <- c(grid_z, s_vals * e1[3] + t * e2[3], NA)
}
for (s in g_lines) {
  grid_x <- c(grid_x, s * e1[1] + t_vals * e2[1], NA)
  grid_y <- c(grid_y, s * e1[2] + t_vals * e2[2], NA)
  grid_z <- c(grid_z, s * e1[3] + t_vals * e2[3], NA)
}

# ============================================================
#  VECTOR GEOMETRY SCALING
# ============================================================
sc      <- 7  
u_vec   <- e1 * sc
v_vec   <- e2 * sc
sum_vec <- u_vec + v_vec

sum_len <- sqrt(sum(sum_vec^2))
sum_dir <- if(sum_len > 1e-10) sum_vec / sum_len else c(0,0,0)

eq_str <- "x + y + z = 0"

# ============================================================
#  PIPELINE LAYER COMPOSITION
# ============================================================
fig <- plot_ly(height = 800) %>%
  config(
    responsive = TRUE,
    displayModeBar = FALSE
  )

# 1. Background Coordinate Framework
fig <- fig %>% add_trace(
  type = "scatter3d", mode = "lines",
  x = axes_lines$x, y = axes_lines$y, z = axes_lines$z,
  line = list(color = "#dfe6e9", width = 2.5),
  showlegend = FALSE, hoverinfo = "none"
)

# 2. Background Structural Ticks
fig <- fig %>% add_trace(
  type = "scatter3d", mode = "lines",
  x = tk_x, y = tk_y, z = tk_z,
  line = list(color = "#dfe6e9", width = 1.5),
  showlegend = FALSE, hoverinfo = "none"
)

# 3. Dynamic Text Annotations for Standard Axis Terminals
fig <- fig %>% add_trace(
  type = "scatter3d", mode = "text",
  x = c(22, 0, 0), y = c(0, 11, 0), z = c(0, 0, 11),
  text = c("<b>X</b>", "<b>Y</b>", "<b>Z</b>"),
  textfont = list(size = 12, color = "#b2bec3", family = "sans-serif"),
  showlegend = FALSE, hoverinfo = "none"
)

# 4. Continuous Subspace Plane Sheet Surface
fig <- fig %>% add_surface(
  x = Xm, y = Ym, z = Zm,
  colorscale  = list(c(0, "#74b9ff"), c(1, "#0984e3")),
  opacity     = 0.45,
  showscale   = FALSE,
  contours    = list(x=list(show=FALSE), y=list(show=FALSE), z=list(show=FALSE)),
  name        = "Subspace Plane (P)",
  hovertemplate = paste0("Plane: ", eq_str, "<extra></extra>")
)

# 5. Overlay Grid Lines on Surface
fig <- fig %>% add_trace(
  type = "scatter3d", mode = "lines",
  x = grid_x, y = grid_y, z = grid_z,
  line        = list(color = "#232F34", width = 2),
  opacity     = 0.15,
  showlegend  = FALSE, hoverinfo = "none"
)

# 6. Parallelogram Extension Line 1
fig <- fig %>% add_trace(
  type = "scatter3d", mode = "lines",
  x = c(u_vec[1], sum_vec[1]), y = c(u_vec[2], sum_vec[2]), z = c(u_vec[3], sum_vec[3]),
  line = list(color = "#00b894", width = 2.5, dash = "dash"),
  showlegend = FALSE, hoverinfo = "none"
)

# 7. Parallelogram Extension Line 2
fig <- fig %>% add_trace(
  type = "scatter3d", mode = "lines",
  x = c(v_vec[1], sum_vec[1]), y = c(v_vec[2], sum_vec[2]), z = c(v_vec[3], sum_vec[3]),
  line = list(color = "#232F34", width = 2.5, dash = "dash"),
  showlegend = FALSE, hoverinfo = "none"
)

# 8. Plane Vector u
fig <- fig %>% add_trace(
  type = "scatter3d", mode = "lines",
  x = c(0, u_vec[1]), y = c(0, u_vec[2]), z = c(0, u_vec[3]),
  line = list(color = "#232F34", width = 5.5),
  showlegend = FALSE, hovertemplate = "u<extra></extra>"
)

# 9. Plane Vector u Arrowhead Cone
fig <- fig %>% add_trace(
  type = "cone", 
  x = u_vec[1], y = u_vec[2], z = u_vec[3], 
  u = e1[1], v = e1[2], w = e1[3],
  sizemode = "absolute", sizeref = 1.2, showscale = FALSE,
  anchor = "tail",
  colorscale = list(c(0, "#232F34"), c(1, "#232F34")),
  hoverinfo = "none", showlegend = FALSE
)

# 10. Plane Vector v
fig <- fig %>% add_trace(
  type = "scatter3d", mode = "lines",
  x = c(0, v_vec[1]), y = c(0, v_vec[2]), z = c(0, v_vec[3]),
  line = list(color = "#00b894", width = 5.5),
  showlegend = FALSE, hovertemplate = "v<extra></extra>"
)

# 11. Plane Vector v Arrowhead Cone
fig <- fig %>% add_trace(
  type = "cone", 
  x = v_vec[1], y = v_vec[2], z = v_vec[3], 
  u = e2[1], v = e2[2], w = e2[3],
  sizemode = "absolute", sizeref = 1.2, showscale = FALSE,
  anchor = "tail",
  colorscale = list(c(0, "#00b894"), c(1, "#00b894")),
  hoverinfo = "none", showlegend = FALSE
)

# 12. Resultant Vector Line: u + v
fig <- fig %>% add_trace(
  type = "scatter3d", mode = "lines",
  x = c(0, sum_vec[1]), y = c(0, sum_vec[2]), z = c(0, sum_vec[3]),
  line = list(color = "#e17055", width = 6.5),
  showlegend = FALSE, hovertemplate = "u + v<extra></extra>"
)

# 13. Resultant Vector u + v Arrowhead Cone
fig <- fig %>% add_trace(
  type = "cone", 
  x = sum_vec[1], y = sum_vec[2], z = sum_vec[3], 
  u = sum_dir[1], v = sum_dir[2], w = sum_dir[3],
  sizemode = "absolute", sizeref = 1.4, showscale = FALSE,
  anchor = "tail",
  colorscale = list(c(0, "#e17055"), c(1, "#e17055")),
  hoverinfo = "none", showlegend = FALSE
)

# 14. Vector Typography Identifiers
fig <- fig %>% add_trace(
  type = "scatter3d", mode = "text",
  x = c(u_vec[1]*1.25, v_vec[1]*1.25, sum_vec[1]*1.2), 
  y = c(u_vec[2]*1.25, v_vec[2]*1.25, sum_vec[2]*1.2), 
  z = c(u_vec[3]*1.25, v_vec[3]*1.25, sum_vec[3]*1.2),
  text = c("<b>u</b>", "<b>v</b>", "<b>u + v</b>"),
  textfont = list(size = 14, color = c("#232F34", "#00b894", "#e17055"), family = "serif"),
  showlegend = FALSE, hoverinfo = "none"
)

# 15. Origin Anchor Node
fig <- fig %>% add_trace(
  type = "scatter3d", mode = "markers",
  x = 0, y = 0, z = 0,
  marker = list(color = "#2d3436", size = 6, symbol = "circle",
                line = list(color = "white", width = 1.5)),
  showlegend = FALSE,
  hovertemplate = "<b>Subspace Origin</b><br>Coordinates: (0,0,0)<extra></extra>"
)

# ============================================================
#  TEXTBOOK DESIGN SCHEME LAYOUT (With Pronounced Walls)
# ============================================================
fig <- fig %>% layout(
  paper_bgcolor = "white",
  margin = list(l = 20, r = 20, t = 20, b = 20),
  scene = list(
    xaxis = list(
      visible = TRUE,
      title = list(text = "x"),
      range = c(-20, 20),
      showgrid = TRUE,
      gridcolor = "#e8e8e8",
      zeroline = TRUE,
      zerolinecolor = "#b0b0b0",
      zerolinewidth = 2,
      # Pronounce the wall background
      showbackground = TRUE,
      backgroundcolor = "#fcfcfc"
    ),
    yaxis = list(
      visible = TRUE,
      title = list(text = "y"),
      range = c(-10, 10),
      showgrid = TRUE,
      gridcolor = "#e8e8e8",
      zeroline = TRUE,
      zerolinecolor = "#b0b0b0",
      zerolinewidth = 2,
      # Pronounce the wall background
      showbackground = TRUE,
      backgroundcolor = "#fafafa"
    ),
    zaxis = list(
      visible = TRUE,
      title = list(text = "z"),
      range = c(-10, 10), 
      showgrid = TRUE,
      gridcolor = "#e8e8e8",
      zeroline = TRUE,
      zerolinecolor = "#b0b0b0",
      zerolinewidth = 2,
      # Pronounce the wall background
      showbackground = TRUE,
      backgroundcolor = "#f5f5f5"
    ),
    aspectmode = "manual",
    aspectratio = list(x = 2, y = 1, z = 1), 
    camera = list(
      eye = list(x = 1.85, y = -1.85, z = 1.35),
      up = list(x = 0, y = 0, z = 1),
      projection = list(type = "orthographic")
    ),
    bgcolor = "white"
  )
)

fig
```

**1. Contains the zero vector**

Substituting $(0,0,0)$ into the plane equation gives

$$\alpha(0) + \beta(0) + \gamma(0) = 0,$$

So the zero vector is in the plane.

**2. Closed under addition**

Let $\mathbf{u} = (x_1, y_1, z_1) \in P$ and $\mathbf{v} = (x_2, y_2, z_2) \in P$. Their sum is defined component-wise as:

$$\mathbf{u} + \mathbf{v} = (x_1 + x_2, \; y_1 + y_2, \; z_1 + z_2)$$

Since both $\mathbf{u}$ and $\mathbf{v}$ lie in the plane $P$, their coordinates must satisfy the plane's defining equation:

$$\alpha x_1 + \beta y_1 + \gamma z_1 = 0$$

and

$$\alpha x_2 + \beta y_2 + \gamma z_2 = 0$$

Adding these two equations yields:

$$\alpha(x_1 + x_2) + \beta(y_1 + y_2) + \gamma(z_1 + z_2) = 0$$

Hence,

$$(x_1 + x_2, \; y_1 + y_2, \; z_1 + z_2) \in P$$

$$\mathbf{u} + \mathbf{v} \in P$$

**3. Closed under scalar multiplication**

Let $\mathbf{u}(x_1, y_1, z_1) \in P$ and let $\lambda \in \mathbb{R}$. Since

$$\alpha x_1 + \beta y_1 + \gamma z_1 = 0,$$

multiplying by $\lambda$ gives

$$\alpha(\lambda x_1) + \beta(\lambda y_1) + \gamma(\lambda z_1) = 0$$

Therefore,$$(\lambda x_1, \lambda y_1, \lambda z_1) \in P$$

$$\lambda\mathbf{u} \in P$$

Since all three conditions are satisfied, every plane through the origin is a subspace of $\mathbb{R}^3$.

Moreover, the subspaces of $\mathbb{R}^3$ are precisely:

-   The trivial subspace $\{(0,0,0)\}$
-   All lines through the origin
-   All planes through the origin
-   $\mathbb{R}^3$ itself
:::

## Vector Space in $\mathbb{R}^n$

We can represent vectors as a sequence of entries (or elements) arranged in a column, called a **column vector**. For example, we write a vector $\mathbf{u}$ in $\mathbb{R}^n$ as

$$\mathbf{u} = \begin{bmatrix} u_1 \\ u_2 \\ \vdots \\ u_n \end{bmatrix}$$

Because writing vertically takes up a lot of space, this exact same vector is often written horizontally using parentheses and commas:

$$\mathbf{u} = (u_1, u_2, \dots, u_n)$$

 

#### Example: Vector $u = (u_1, u_2)$

For example, a vector $\mathbf{u}$ in a two-dimensional space ($\mathbb{R}^2$) can be expressed algebraically as an ordered pair of its entries, $u_1 = 1.5$ and $u_2 = 1$, written as $\mathbf{u} = (1.5, 1)$. Geometrically, it is represented as an arrow, or directed line segment, starting at the origin in the coordinate plane and ending at the point with coordinates $u_1$ and $u_2$ (@fig-vector).

```{r}
#| echo: false
#| warning: false
#| fig-align: center
#| label: fig-vector
#| fig-cap: 'In a Cartesian coordinate system, a vector in $\mathbb{R}^2$ is geometrically represented as a directed line segment (an arrow).'


library(ggplot2)

# --- 1. Define Vector and Label Data ---
u1 <- 1.5
u2 <- 1.0

point_data <- data.frame(
  x = u1,
  y = u2,
  label = "u == (list(1.5, 1))"
)

# --- 2. Build the Plot ---
ggplot() +
  # Clear theme to remove background grid lines completely
  theme_void() + 
  theme(
    plot.margin = margin(20, 20, 20, 20)
  ) +
  
  # 1. Vertical and Horizontal Dashed Projection Lines (Palette from Image 1)
  geom_segment(aes(x = u1, y = u2, xend = u1, yend = 0), 
               color = "#555555", lty = "dashed", linewidth = 0.5) +
  geom_segment(aes(x = u1, y = u2, xend = 0, yend = u2), 
               color = "#555555", lty = "dashed", linewidth = 0.5) +

  
  # 2. Axis Lines (x and y) using the gray color (#555555)
  geom_segment(aes(x = 0, y = 0, xend = 2.0, yend = 0), 
               arrow = arrow(length = unit(0.25, "cm"), type = "open"), 
               color = "#555555", linewidth = 0.5) +
  geom_segment(aes(x = 0, y = 0, xend = 0, yend = 1.8), 
               arrow = arrow(length = unit(0.25, "cm"), type = "open"), 
               color = "#555555", linewidth = 0.5) +
  
  # Axis Labels (Muted gray text)
  annotate("text", x = 2.15, y = 0, label = "x", fontface = "italic", color = "#555555", size = 6) +
  annotate("text", x = 0, y = 1.95, label = "y", fontface = "italic", color = "#555555", size = 6) +
  annotate("text", x = -0.05, y = -0.08, label = "O (0,0)", fontface = "italic", color = "#555555", size = 5) +
  
  # Axis Ticks / Intercept Labels for the components
  annotate("text", x = u1, y = -0.1, label = "u[1] == 1.5", parse = TRUE, color = "#555555", size = 5) +
  annotate("text", x = -0.17, y = u2, label = "u[2] == 1", parse = TRUE, color = "#555555", size = 5) +
  
  # 3. The Vector Arrow using the Dark Red outline color (#8B0000)
  geom_segment(aes(x = 0, y = 0, xend = u1, yend = u2), 
               arrow = arrow(length = unit(0.35, "cm"), type = "closed"), 
               color = "#8B0000", linewidth = 1.0) +
  
  # Vector Name Label
  annotate("text", x = u1 * 0.45, y = u2 * 0.55 + 0.1, label = "u", fontface = "bold", color = "#8B0000", size = 5) +
  
  
  # 5. Math Coordinate Label matching the dark red style
  geom_text(data = point_data, aes(x = x, y = y, label = label), 
            color = "#8B0000", parse = TRUE, size = 5, hjust = -0.15, vjust = -0.15) +
  
  # 6. Fix coordinate ratio and set plot boundaries
  coord_fixed(xlim = c(-0.3, 2.5), ylim = c(-0.3, 2.1))
```

##### *Implementation in R*

We can initialize our specific vector $\mathbf{u} = (1.5, 1)$ and print its entries to the console as follows:

```{r}
u <- c(1.5, 1)
u
```

## Vector Addition in $\mathbb{R}^n$

If $\mathbf{a} = (a_1, a_2, \dots, a_n)$ and $\mathbf{b} = (b_1, b_2, \dots, b_n)$ are two vectors in $\mathbb{R}^n$, then the sum of $\mathbf{a}$ and $\mathbf{b}$ is the vector:$$\mathbf{a} + \mathbf{b} = (a_1 + b_1, a_2 + b_2, \dots, a_n + b_n)$$

We add two vectors, $\mathbf{u} = (u_1, u_2)$ and $\mathbf{v} = (v_1, v_2)$, by adding their components:

$$u + v = (u_1, u_2) + (v_1, v_2) = (u_1 + v_1, u_2 + v_2)$$

```{r}
#| echo: false
#| warning: false
#| fig-align: center
#| label: fig-vector2
#| fig-cap: 'In a Cartesian coordinate system, a vector in $\mathbb{R}^2$ is geometrically represented as a directed line segment (an arrow).'


# 1. Define vector components
ux <- 3; uy <- 1
vx <- 1; vy <- 2
rx <- ux + vx  # 4
ry <- uy + vy  # 3

# 2. Build the blank theme canvas
ggplot() +
  # Custom Axis Lines (Charcoal Gray)
  geom_segment(aes(x = -0.5, y = 0, xend = 4.8, yend = 0), 
               arrow = arrow(length = unit(0.2, "cm"), type = "open"), color = "#555555", linewidth = 0.6) +
  geom_segment(aes(x = 0, y = -0.5, xend = 0, yend = 3.8), 
               arrow = arrow(length = unit(0.2, "cm"), type = "open"), color = "#555555", linewidth = 0.6) +
  
  # Axis Labels
  annotate("text", x = 4.7, y = -0.2, label = "x", fontface = "italic", size = 4.5, color = "#555555") +
  annotate("text", x = -0.2, y = 3.7, label = "y", fontface = "italic", size = 4.5, color = "#555555") +
  annotate("text", x = -0.15, y = -0.15, label = "O", fontface = "italic", size = 4, color = "#555555") +
  
  # Parallelogram / Projection Guidelines (Dashed Lines)
  geom_segment(aes(x = ux, y = uy, xend = rx, yend = ry), lty = "dashed", color = "#777777", linewidth = 0.5) +
  geom_segment(aes(x = vx, y = vy, xend = rx, yend = ry), lty = "dashed", color = "#777777", linewidth = 0.5) +
  
  # Vector 'u' (Blue)
  geom_segment(aes(x = 0, y = 0, xend = ux, yend = uy), 
               arrow = arrow(length = unit(0.25, "cm"), type = "closed"), color = "#1f77b4", linewidth = 1.0) +
  # Vector 'v' (Green)
  geom_segment(aes(x = 0, y = 0, xend = vx, yend = vy), 
               arrow = arrow(length = unit(0.25, "cm"), type = "closed"), color = "#2ca02c", linewidth = 1.0) +
  # Resultant Vector 'u + v' (Maroon/Red)
  geom_segment(aes(x = 0, y = 0, xend = rx, yend = ry), 
               arrow = arrow(length = unit(0.3, "cm"), type = "closed"), color = "#8B0000", linewidth = 1.2) +
  
  # Mathematical Labels (Parsed textmath expressions)
  annotate("text", x = 1.7, y = 0.3, label = "u", fontface = "bold.italic", size = 5, color = "#1f77b4") +
  annotate("text", x = 0.3, y = 1.2, label = "v", fontface = "bold.italic", size = 5, color = "#2ca02c") +
  annotate("text", x = 1.8, y = 1.8, label = "u + v", fontface = "bold.italic", size = 5, color = "#8B0000") +
  
  # Terminal Coordinate Labels
  annotate("text", x = ux + 0.4, y = uy, label = "u == (list(3, 1))", parse = TRUE, size = 4, color = "#1f77b4") +
  annotate("text", x = vx - 0.4, y = vy + 0.1, label = "v == (list(1, 2))", parse = TRUE, size = 4, color = "#2ca02c") +
  annotate("text", x = rx + 0.7, y = ry, label = "w == (list(4, 3))", parse = TRUE, size = 4.2, color = "#8B0000") +
  
  # Force layout constraints matching original graphic style
  coord_fixed() +
  theme_void()
```

 

#### Example: Vector addition u + v

$$w = u + v = (3, 1) + (1, 2) = (3 + 1, 1 + 2) = (4, 3)$$

##### *Implementation in R*

```{r}
u <- c(3, 1)
v <- c(1, 2)

w <- u + v
w
```
